---
title: "Does Not Playing Hockey Make You Worse At Hockey? Investigating the Impact of the COVID-19 Pandemic on Hockey Player development"

# to produce blinded version set to 1
blinded: 1

authors: 
- name: Michele Sezgin
  thanks: The authors gratefully acknowledge Sam Ventura, Ron Yurko, Ben Baumer, and Ryne Vankrevelen
  affiliation: Department of Statistical and Data Sciences, Smith College
  
- name: Jackie Jovanovic
  affiliation: Department of Statistics, Elon University

keywords:
- sports analytics
- causal inference
- Coronavirus pandemic

abstract: |
  During 2020-2021, many hockey leagues had shortened seasons or no seasons due to the restrictions on play caused by the COVID-19 pandemic. The Ontario Hockey League did not play any games during this “COVID” season, which raises the question of whether not playing games during the 2020-2021 COVID season negatively impacted player development (or even caused players to get worse). Statistical and causal inference methods were employed to assess the impact of not playing during the COVID season, including Ordinary Least Squares, Gamma, Mixed Effects, and matched regression models, as well as a Bayesian Additive Regression Model. All models pointed to no significant difference in post-COVID performance for treated (played during COVID season) vs control (did not play during COVID season) groups. Future research should focus on studying other populations, experimenting with alternative variable definitions, and studying player development curves to more thoroughly assess treatment impact.

bibliography: bibliography.bib
output: rticles::asa_article
---

```{r, packages, include = FALSE}
# load required packages
library(cowplot)
library(tidyverse)
library(kableExtra)
options(warn=-1)
```

```{r, data, include = FALSE}
ohl_filtered <- read_csv("../data/ohl_filtered.csv")
```

```{r, rds-objects, include = FALSE}
# Data
data_integrity <- readRDS("./supplementary-materials/rds-files/data_integ_table.rds")
# EDA plots
prelim_plots <- readRDS("./supplementary-materials/rds-files/prelim_plots.rds")
age_plots <- readRDS("./supplementary-materials/rds-files/age_plots.rds")
gp_plots <- readRDS("./supplementary-materials/rds-files/gp_plots.rds")
ppp_plots <- readRDS("./supplementary-materials/rds-files/ppp_plots.rds")
pos_plots <- readRDS("./supplementary-materials/rds-files/pos_plots.rds")
# Results
rmse_table <- readRDS("./supplementary-materials/rds-files/rmse_table.rds")
ols_plot <- readRDS("./supplementary-materials/rds-files/ols_plot.rds")
shane_demo <- readRDS("./supplementary-materials/rds-files/shane_demo.rds")
bart_plots <- readRDS("./supplementary-materials/rds-files/bart_plots.rds")
bart_cates <- readRDS("./supplementary-materials/rds-files/bart_cates.rds")
# Limitations
ols_lim_plots <- readRDS("./supplementary-materials/rds-files/ols_lim_plots.rds")
```


# Introduction

During 2020-2021, many hockey leagues had shortened seasons or no seasons due to the restrictions on play caused by the COVID-19 pandemic. Among these leagues were National Hockey League (NHL) feeder leagues, such as the Ontario Hockey League (OHL), the Western Hockey League, and the Quebec Major Junior Hockey League. Some players experiencing restrictions on play participated in other leagues or tournaments during the 2020-21 season, while others did not. This raises the question of whether not playing games during the 2020-2021 COVID season negatively impacted player development (or even caused players to get worse).

To answer this question, we examined data from the OHL, which did not play any games during the 2020-21 season due to the state of the COVID-19 pandemic (1). Some players from this league found leagues abroad (U.S. or Europe) to play in, while others recorded no games during 2020-2021, which we will call the COVID season. While some leagues may have played a partial season, OHL players either played in another league, or they did not play at all. This provided a natural experiment, allowing us to not only use traditional statistical inference but also causal inference methods. 

There has been much speculation from coaches, scouts, sports commentators, and fans about the effects of the COVID season on player progression \cite{sportsnet, traikos}.  The question many asked prior to the COVID season remains (at least quantitatively) unanswered: “what will be the impact on athletes unable to compete for well over a year during a key stretch of their development?” @sportsnet. Some coaches theorized that players would become better and stronger having had time off to reevaluate their individual performance and work on more skill, speed, and strength training @sportsnet, while others feared the lack of game play may have dulled the sharpening hockey skills and IQ of developing players @traikos. There have been plenty of skaters who had to take time off from their hockey career due to injury and came back better and stronger than before \cite{dixon, kreiser}. However, these are typically NHL players with access to extensive training and rehabilitation resources, something that not all OHL players may have had. Additionally, these players are typically not taking time off during pivotal developmental years; they are already well-established professional players.

# Previous Work

Previous work has focused on psychological aspects in player development, such as mutual trust and respect between athletes and their coaches, or frameworks through which players develop \cite{lefebvre, soberlak}. But little research was found on pivotal developmental years or typical development curves for ice hockey players, though some work has been done to investigate developmental curves in swimming @born. This study pointed to significant perormance growth from ages 10 to roughly 17 depending on the group analyzed, which could potentially translate to hockey development and OHL players. The prior research most similar to our research investigated the performance impacts of concussions in the NHL \cite{buckley, neustadtl}. Neustadtl et al. found no significant differences in pre- and post-concussion performance for the 269 NHL players studied. Buckley at al. again found no significant difference in pre and post- performance for players who missed playing time due to injury and those who missed playing time due to non-injury reasons. However, no research was found investigating the impact of reduced playing time in developing (major junior) hockey players.

# Data

Our dataset includes skaters (excludes goalies) who played in the Ontario Hockey League (OHL) during both the pre-COVID (2019-2020) and post-COVID (2021-2022) seasons. The dataset also contains information regarding other leagues they have participated in during their career. There is data about season, team, league, points, games played, position, and drafted status. Each row is a player on a certain team in a specific season. There are 219 observations in the dataset. The data was sourced from Elite Prospects and was supplied by our external advisor.

```{r, sample-rows, echo = FALSE}
ohl_filtered %>% 
  dplyr::select(first_name, last_name, position, ppg_19,
                gp_21, age_continuous, treatment, ppg_21) %>%
  filter(first_name %in% c("Shane","Brandt", "Logan", "Wyatt") & last_name %in% c("Wright", "Clarke", "Mailloux", "Johnston")) %>%
  arrange(desc(last_name)) %>%
  knitr::kable(format = "latex", caption = "Sample data rows and variables") %>%
  kable_styling(latex_options="scale_down")
```

## Added Variables

If a skater played for several teams during one season, including the COVID season, the team for which they played the most games was selected. Statistics for all other teams were dropped. This was done to control for team effect on performance.

\(\hspace{-0.25in}\)**Player performance**: To approximate player performance in the post-COVID season, we calculated each player’s points per game (PPG) as PPG = (goals + assists)/games played in the post-COVID season. 

\(\hspace{-0.25in}\)**Previous player performance**: To approximate player performance in the pre-COVID season, we computed each player’s PPG in the pre-COVID season.

\(\hspace{-0.25in}\)**Treatment**: We defined a player as treated if they participated in at least one game during the post-COVID season. This allowed us to retain players who recorded relatively few games in elite competitions like the International Ice Hockey Federation World Junior Championship as treated. Thus ensuring we were not obscuring the effects of these competitions on development. There are 67 treated players and 152 control players.

\(\hspace{-0.25in}\)**Age**: We defined age as the continuous age of the player as of January 1st, 2020.
Previous Player Performance : To control for player performance in the pre-COVID season, we approximated player performance using player PPG in the pre-COVID season.

\(\hspace{-0.25in}\)**Drafted**: Whether a player was drafted in 2020 or earlier
Relative PM : Relative plus-minus (PM) is defined as a player's PM relative to the average PM of their team.

\(\hspace{-0.25in}\)**Ranked PM**: Ranked plus-minus defines how a player’s PM ranks among those of their teammates. If there are n players on a team, each player will receive a PM ranking 1-n.

## Data Integrity

Through analysis and comparison with OHL records, we estimate that we have 99% of the data from the 2019-2020 and 2021-2022 regular seasons (Table 3).

```{r, data-integrity, echo = FALSE}
data_integrity %>%
  knitr::kable(format = "latex", caption = "Missing goals for each OHL team in the 2019-2020 and 2021-2022 regular seasons") %>%
  kable_styling(latex_options="scale_down")
```

## EDA

### Preliminary Visualizations

The distribution of player performance is right-skewed, showing that most players do not generate many points per game. The preliminary hypothesis that not playing during the COVID season negatively impacted player performance is motivated by the plot of player performance distributions conditioned on treatment. It appears that players who played during the COVID season performed better in the post-COVID season, which motivated our analysis. Our models attempt to tease out if this difference is “real”. \(\\\)

```{r, prelim-plots, echo = FALSE}
prelim_plots
```
\(\hspace*{\fill} \footnotesize \text{Figure 1. preliminary visualizations}\)

We investigated relationships between potential confounders and our treatment and response to aid in variable selection. The variables we hypothesized as potential confounders are age, games played, drafted status, previous player performance, and position.

### Age

We hypothesized that older players would be more likely to have better player performance, because they are further along in their development. Interestingly, the below plots do not support a strong relationship between age and player performance. We additionally hypothesized that older players would be more likely to be treated using the same logic as before. Better players are more likely to be selected to play in leagues, and older players tend to be better. However, the distributions do not differ significantly when conditioning on treatment. \(\\\)

```{r, age-plots, echo = FALSE}
age_plots
```
\(\hspace*{\fill} \footnotesize \text{Figure 2. age vs. treatment and response}\)

### Games Played

The data supports that players who play more games are likely to be higher performing players. I.e. “good players don’t get benched”, as expected. However, there are no overwhelming differences in the distributions of player performance when conditioning on treatment, and the instability in the distribution of player performance for players who played during the COVID season is likely due to the smaller sample size.

```{r, gp-plots, echo = FALSE}
gp_plots
```
\(\hspace*{\fill} \footnotesize \text{Figure 3. games played vs. treatment and response}\)

### Previous Player Performance

We hypothesized that higher performing players pre-COVID would likely be higher performing players post-COVID, because good players tend to keep being good. The data supports that there is a positive, moderate linear relationship between previous player performance and post-COVID player performance. We additionally thought that higher performing players pre-COVID would be more likely to be treated, because it would be easier to convince overseas leagues to take on better players, a notion that appears to be supported by the data.

```{r, ppp-plots, echo = FALSE}
ppp_plots
```
\(\hspace*{\fill} \footnotesize \text{Figure 4. previous player performance vs. treatment and response}\)

### Position

Lastly, we know that defensemen will be lower performing players because of the way we have defined player performance and because of the nature of the position, a notion that is supported by the data. However, there is no reason to believe that position would influence probability of treatment, which is again supported by the data. 

```{r, pos-plots, echo = FALSE}
pos_plots
```
\(\hspace*{\fill} \footnotesize \text{Figure 5. position vs. treatment and response}\)

# Methods

Before performing causal analysis, we used traditional statistical inference methods to determine whether playing during the COVID season had a significant effect on post-COVID player performance while controlling for variables (and interactions between variables) suspected to be associated with the response through exploratory data analysis (EDA). When possible, we attempted a few different measures of player performance, including PPG z-scores (to control for differences in overall scoring across seasons) and relative plus-minus scores. All code is available in the supplementary code document of the supplementary materials folder @author.

For all methods \(\epsilon \sim N(0, \sigma^2)\)

## Regression

### OLS

We fit an Ordinary Least Squares (OLS) regression model without interaction, because it is the simplest and most interpretable model to assess the significance of playing during COVID, while still controlling for variables that could be confounding our analysis.

\vspace{-.5cm}

\[\text{Player Performance} = \beta_0 + \beta_1\text{Treatment} + \beta_2\text{Forward} +  \beta_3\text{Previous Performance} \hspace{.1cm} + \]
\[\beta_4\text{Games Played Post-COVID} + \beta_5\text{Age} + \epsilon\]

### Interaction OLS

We fit an OLS model with interaction to account for relationships observed between explanatory variables during the EDA process. We also tested whether the additional complexity of this full model significantly increased predictive power over the nested model.

\vspace{-.5cm}

\[\text{Player Performance} = \beta_{0} + \beta_{1}\text{Treatment} + \beta_{2}\text{Forward} +  \beta_{3}\text{Previous Performance} \hspace{.1cm} +\] 
\[\beta_{4}\text{Games Played Post-COVID} + \beta_{5}\text{Age} + \beta_{6}\text{Position*Previous Performance} \hspace{.1cm} + \]
\[\beta_{7}\text{Age*Previous Performance} + \beta_{8}\text{Previous Performance*Treatment} + \beta_{9}\text{Games Played} + \epsilon\]

### Gamma

PPG is right skewed and bounded between 0 and some positive number, so we believed PPG may be Gamma-distributed. We performed simple and interaction Gamma regression with the log link function in an attempt to more accurately model the relationship between our response and explanatory variables.

\vspace{-.5cm}

- \(\text{Player Performance} \sim \text{Gamma}(\alpha, \lambda)\)
- \(\text{for the }i^{th} \text{player, } i= 1,..., 219,\)
\[E[\text{Player Performance}_{i}] =  e^{\beta_0} \text{Treatment}^{\beta_1}  \text{Forward}^{\beta_2} \text{Previous Performance}^{\beta_3}\]
\[\text{Games Played Post-COVID}^{\beta_3} \text{Age}^{\beta_5}\]

### Mixed-effects Model

Lastly, we fit a mixed-effects model to determine if the effect of playing during the COVID season was significantly different across leagues. The slope and intercept of the regression line varied according to the league. Players who did not participate during the COVID season were grouped into a “league” called “NONE”.

**Level One**:

- \(\text{for the }i^{th} \text{player, } i= 1,..., 219,\)

\[Y_i = a_i + b_i\text{Treatment} + \epsilon_i\]

- \(\epsilon_i \sim N(0, \sigma^2)\)

**Level Two**:

- \(a_i = \alpha_0 + \alpha_1\text{Forward} + \alpha_2\text{Previous Performance} +  \alpha_3\text{Games Played Post-COVID} + \alpha_4\text{Age} + u_i\)

- \(b_i = \beta_0 + \beta_1\text{Forward} +  \beta_1\text{Previous Performance} + 
\beta_3\text{Games Played Post-COVID} + \beta_4\text{Age} + v_i\)

\begin{align*}
\begin{bmatrix}u_{i}\\
v_{i}
\end{bmatrix} &\sim  N
\begin{pmatrix}
\begin{bmatrix}
0\\
0
\end{bmatrix}\!\!,&
\begin{bmatrix}
\sigma_{u}^2 & \\
\rho\sigma_{u}\sigma_{v} & \sigma_{v}^2
\end{bmatrix}
\end{pmatrix}
\end{align*}

## Causal Methods

### Matching

Matching on covariates allowed us to better control for confounding variables and more accurately estimate the true impact of the treatment. The results fit on matched data were compared to the results of the model fit on unmatched data. Control and treatment observations were matched on  games played post-COVID, position, points pre-COVID, previous player performance, age, and ranked PM pre-COVID. "Optimal" matching method and GAM measure of distance were used, and a 1-to-1 matching ratio was used due to the small sample size. The model is the same as the OLS model, but fit with matched data.

### BART

An appeal of the Bayesian Additive Regression Trees (BART) model is that it controls overfitting, which is one issue with normal additive regression trees. Each tree tries to account for something new in the model. Once all added, accurate estimates are produced. This allows for causal inferences to be drawn. The workflow used is taken from Joshua Bon's vignette on using the tidytreatment package with BART. 

A variable selection model was used to find important variables for obtaining propensity scores. The variable selection model was a continuous BART model regressing all covariates except for the treatment against post-COVID player performance. The burn-in period was set to 2000 iterations, with posterior draws set to 5000. Variables were deemed important if their average inclusion in the model was equal to or greater than 50%. The most important variables were used to produce propensity scores using a probit BART model, with a burn-in of 2000 draws and 5000 posterior draws. Finally, all covariates, including treatment, and propensity scores were used in the final BART model to predict post-COVID player performance. The final model produced a posterior distribution of estimated post-COVID player performance for each player, with a burn-in of 10,000 draws and 200 posterior draws per player. The method of obtaining the posterior is outlined by @chipman.

For all models "let \(T\) denote a binary tree consisting of a set of interior node decision rules
and a set of terminal nodes, and let \(M = \{\mu_1, \mu_2, . . . ,\mu_b\}\) denote a set of
parameter values associated with each of the \(b\) terminal nodes of \(T\)" @chipman.

**Variable selection model**:

\[\text{Player Performance} = \sum_{j=1}^{25} g(x;T_{j}, M_{j}) + \epsilon\]

- \(x = \{ \text{Games Played Pre-COVID}, \text{Games Played Post-COVID}, \text{Points Pre-COVID}, \text{Age}, \\ \text{Ranked PM Pre-COVID}, \text{Relative PM Pre-COVID}, \text{PM Pre-COVID}, \text{Position}, \\
\text{Previous Player Performance} \}\)

- \(x_{important}\) are the most variables in this model, deemed important if average inclusion was greater than or equal to 50%.

**Propensity score model**:

\vspace{-.5cm}

\[\text{Propensity Score} = P(\text{Treatment} = \text{played } | \hspace{.1cm} x_{important}) = \Phi(\sum_{j=1}^{200} g(x_{important};T_{j}, M_{j}) + \epsilon)\]

- \(\Phi\) is the cumulative distribution function of the standard normal distribution.

**Final Predictive Model**:

\[ \text{Player Performance} = \sum_{j=1}^{200} g(x, \text{Treatment}, \text{Propensity Score};T_{j}, M_{j}) + \epsilon\]

# Results

Across all models there was no evidence of a treatment effect given the observed data, regardless of the measure of player performance chosen (PPG, PPG z-scores, ranked PM, relative PM). PPG was ultimately chosen as the response measure because of its interpretability. The root mean squared error (RMSE) values for each of the models according to ascending error is summarized in Table 3.

\begin{table}

\caption{\label{Table 3.}RMSE values for all models}
\centering
\begin{tabular}[t]{l|r}
\hline
model & RMSE\\
\hline
BART & 0.2254971\\
\hline
mixed effects & 0.2386779\\
\hline
OLS interaction & 0.2548995\\
\hline
OLS & 0.2553530\\
\hline
intercept-only & 0.2662603\\
\hline
matched & 0.4018033\\
\hline
gamma simple & 1.1395561\\
\hline
gamma interaction & 1.2248592\\
\hline
\end{tabular}
\end{table}


## Regression

The regression model that best balances accuracy and parsimony is the OLS model.

\(\widehat{\text{Player Performance}} = 1.14 + 0.05*\text{Treatment} + 0.18*\text{Forward} +  0.89*\text{Previous Performance} \hspace{.1cm} + .006*\text{Games Played Post-COVID} - .07*\text{Age}\)
\(\vspace{.25cm}\)

```{r, ols-plot, echo = FALSE}
ols_plot
```

- Coefficients for position, games played in post-COVID season, age, and previous player performance were all found to be statistically significant from 0 at the \(\alpha = 0.01\) level.

- The model explains approximately 59% of the variation in player performance in the post-COVID season, which is a statistically significant amount at the \(\alpha = .001\) level \(F(5, 213) = 62.88, p<.001\).

## BART

The variables deemed important in the variable selection model were games played post-COVID, points pre-COVID, relative PM pre-COVID, position, and previous player performance.

\(x_{important} = \{ \text{Games Played Post-COVID}, \text{Points Pre-COVID}, \text{Age}, \\ \text{Relative PM Pre-COVID}, \text{Position}, \text{Previous Player Performance}\}\)
\(\vspace{.25cm}\)

The estimated effects do not differ significantly from the line at 0, which indicates that there is not sufficient evidence of a treatment effect. The histogram of treatment effects is bell-shaped and fairly symmetric. The mean treatment effects show that there could be a slightly positive, though extremely small treatment effect for all subjects, but this is shown to be not significant in Figure 7. 

```{r, bart-plots, echo = FALSE}
bart_plots
```
\(\hspace*{\fill} \footnotesize \text{Figure 6. BART posterior draws and means plots}\)

The confidence intervals for Conditional Average Treatment Effect support the same conclusions. Similar to the mean treatment effects, most of the point estimates in Figure 7 are positive. But since all of the confidence intervals contain 0, this visualization is compatible with no treatment effect.

```{r, bart-cates, echo = FALSE}
bart_cates
```
\(\hspace*{\fill} \footnotesize \text{Figure 7. BART CATE plot}\)

# Discussion

## Interpretations of Models

### OLS Regression Model

Player performance in the post-COVID season is expected to increase by 0.05 PPG on average for skaters who played during COVID, when holding constant position, previous player performance, games played in the post-COVID season, points scored in the pre-COVID season, and age. Over the course of a 68 game season in the OHL, the treatment effect would result in only approximately 3 extra points. In this model, the coefficient for whether someone played during the COVID-season was not statistically significantly different from zero. There is no evidence that playing during the COVID season impacted player performance in the post-COVID season for players who played in both the pre- and post-COVID seasons in the OHL. 

We are 95% confident that the change in player performance in the post-COVID for players who played during the COVID season is between -0.0185 and 0.149 PPG, holding constant position, previous player performance, games played in the post-COVID season, points scored in the pre-COVID season, and continuous age. This model can be used to predict player performance in the post-COVID season based on a variety of factors in the pre-, COVID, and post-COVID seasons among skaters who played in the OHL in both the pre- and post-COVID seasons. For example, using Shane Wright as our observation (Table 4), the model predicts his post-COVID player performance to be 1.61 points per game, while his actual points per game was slightly lower at 1.49.

\begin{table}
\caption{\label{Table 4.}Shane Wright stats}
\centering
\begin{tabular}{l|l|r|r|r|r|r}
\hline
treatment & position & ppg\_19 & gp\_21 & age\_continuous & ppg\_21 & ols\_pred\\
\hline
Played & F & 1.137931 & 63 & 15.98904 & 1.492063 & 1.606924\\
\hline
\end{tabular}
\end{table}

## Limitations

### Model Limitations

#### OLS:

The model conditions for inference were not all met. The relationship between our explanatory variables and post-COVID season PPG looks roughly linear, so the linearity condition of OLS is met. However, the skaters in the dataset are playing together on the same teams in the same league and influencing each other’s player performance. Because we were using player level data, and not play level data, we could not attempt to account for this non-independence. Therefore, this condition is not met. The residuals seem roughly normally distributed, as shown in the normal QQ plot, so this condition is met. There seems to be roughly constant variance in residuals across all levels of our explanatory variables, so the error homogeneity condition is met.

```{r, ols-limitation-plots, echo = FALSE}
ols_lim_plots
```

#### Mixed Effects

Though the mixed effects model shows a significant positive effect for some teams, the confidence intervals were generated using as few as one observation in many cases. Teams where a significant positive effect is shown had at most eight observations. Because of these small sample sizes, we did not believe that our confidence intervals nor our league effect size estimates were credible. Additionally, the model did not converge.

#### Matched regression

Though the the regression model fit on matched data did not support a significant treatment effect, the matching was not ideal. There were not enough observations in the dataset to support extremely precise matching between treatment and control units.

### Variable and Data Limitations

Though we hypothesized that drafted status would have a significant effect on treatment status, unfortunately not enough players in our dataset were drafted prior to the COVID season for this variable to be meaningful (only three players were drafted pre-COVID). We hypothesized that if a player was drafted, it is more likely that they would be able to play during the COVID season, because their NHL team would have the resources to convince a league or team to take the player on for that year. Additionally, we do not have any record of players who left the OHL post-COVID. It is possible that there are a significant number of players who played during the COVID season and became too advanced to continue playing in the OHL, resulting in them leaving to play in other leagues. This missing data could be confounding our analysis, because if these hypotheticals were true, then including these players would likely result in the treatment effect being significantly positive.

Our estimates of player performance were biased towards offensive production. Variables such as time on ice, line number, shots blocked, successful passes, and battles won, among other measures could be combined to create a more comprehensive metric to approximate player performance, which may show a different treatment effect. In general there are many confounding variables that were either not present in our dataset or cannot be easily measured. For example, possible confounders that were not measured in our dataset are time spent practicing per week (on ice, off ice, etc), coaching support, diet, etc.

# Future Work

Future models could utilize weighting and account for dependence of observations. If models were weighted so that players who played more games contributed more to the overall outcome of the model and significance of the treatment, this would likely result in more robust estimates of the treatment effect. Player performance for players who only played a small number of games may not be as accurate as player performance for players with more games. Additionally, robust variances could be used when performing inference for models because of the dependence of our residuals. Play-level data could also be used to address the dependence of residuals in the dataset.

Ideally, we would like to study the effects of the COVID-19 pandemic restrictions and prolonged periods without play on other populations. For example, we would like to investigate if our results generalize to goalies, other major junior leagues, and the NHL/NHL taxi squad. There are other scenarios where players experienced prolonged periods without play. Players may be forced to take time off due to injury, personal reasons, or scenarios like the NHL  lockouts. 

We would also like to facilitate a comparison of typical hockey player development curves compared to COVID-impacted curves, and investigate the impact of our treatment on long-term growth. Even if not playing during the COVID season negatively impacted the development of any population of players, would this impact last forever? Or is it the case that these players would catch up to their counterparts who played during the COVID season in the long-term? As time passes and more data is collected post-COVID, future work could focus on assessing the long-term impact of playing or not playing during the COVID season or not playing during other seasons.

# Conclusion

When studying the effects of the pandemic hockey restrictions on OHL hockey player performance, no significant effect was found between those who played during the pandemic and those who didn't. Both statistical and causal inference methods showed no significant evidence of a treatment effect when controlling for relevant confounding variables, though model and data limitations may have contributed to this null effect. Further research could investigate other populations and variable definitions to determine if our results generalize beyond the population studied.

# Acknowledgments

Thank you to our external advisors, professors, and universities, we could not have completed this project without them and we sincerely appreciate their help.

\pagebreak

<!--to our external advisor Dr. Sam Ventura and our professor and advisor, Dr. Ron Yurko Ben Baumer, Smith College and Dr. Ryne Vankrevelen, Elon University. We could not have completed this project without you all and we sincerely appreciate your help. -->
