---
title: "Meeting 3 Prep"
output: 
  html_document:
    code_folding: hide
date: '2022-07-17'
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = TRUE)
```

```{r, message=FALSE}
# load packages
library(tidyverse)
library(RColorBrewer)
library(ggbeeswarm)
# load data
library(readr)
sams_ohl_data_request <- #read_csv("C:/Users/jacki/Desktop/REU/hockey/sams_ohl_data_request.csv")
read_csv("sams_ohl_data_request.csv")
View(sams_ohl_data_request)
# filter years
recent <- sams_ohl_data_request %>% 
  filter(season %in% c("2019-2020", "2020-2021", "2021-2022"))
# add points per game columns
recent <- recent %>% 
  mutate(ppg = pts/gp)
# more data frames
all3 <- recent %>% 
  filter(season %in% c("2019-2020", "2020-2021", "2021-2022")) %>%
  group_by(league, season) %>%
  summarize(count = n()) %>%
  pivot_wider(names_from = season, values_from = count) %>%
  filter(!is.na(`2019-2020`) & !is.na(`2020-2021`) & !is.na(`2021-2022`))
all_seasons <- recent %>% 
  filter(league %in% c("Belarus Cup", "International-Jr", "MHL",
                       "NOJHL", "Slovakia U20", "Slovakia2",
                       "USHL"))
drafted <- recent %>% 
  filter(!is.na(draft_year))
recent2 <- recent %>% 
  mutate(got_drafted = case_when(!is.na(draft_year) ~ 'Yes',
                                 TRUE ~ 'No'))
```


### Alternatives to boxplots

1. Jitter plot / strip plot
```{r, warning = FALSE}
recent %>% 
  ggplot(aes(x = factor(season), y = ppg)) +
  geom_jitter(alpha = 0.4, aes(color = position),
              position=position_jitter(0.2)) +
  theme_bw()
```

2. Violin
```{r, warning = FALSE}
recent %>% 
  ggplot(aes(x = ppg, y = factor(season))) +
  geom_violin(fill = "cornflowerblue") +
  theme_bw()
```

3. Beeswarm
```{r, warning = FALSE}
recent %>% 
  ggplot(aes(y = ppg)) +
  geom_beeswarm(aes(x = ""), cex = 3, alpha = 0.5) +
  theme_bw() +
  coord_flip()
```

4. Density plot with rugs
```{r, warning = FALSE}
recent %>% 
  ggplot(aes(x = ppg)) +
  geom_density(aes(fill = season), alpha = 0.3) +
  geom_rug(alpha = 0.5) +
  theme_bw() 
```

## Confounding Variables
### Player Quality

### Drafted
#### If drafted or not:
```{r, warning = FALSE}
summary(lm(ppg ~ got_drafted, data = recent2))
recent2 %>% 
  ggplot(aes(x = ppg)) +
  geom_density(aes(fill = got_drafted), alpha = 0.5) +
  theme_bw() +
  labs(x = "points per game", fill = "drafted") +
  facet_wrap(~season)
```

- Very low R-squared value, significant p-value, but we can't use these statistics because model does not meet conditions for inference.

#### Draft pick number / round: 
```{r}
summary(lm(ppg ~ overall_pick_num, data = drafted))
drafted %>% 
  ggplot(aes(x = gp, y = overall_pick_num)) +
  geom_point(alpha = 0.5) +
  geom_smooth() +
  labs(x = "games played", y = "draft pick number") +
  theme_bw()
drafted %>% 
  ggplot(aes(x = factor(round), y = gp)) +
  geom_jitter(alpha = 0.4, aes(color = position),
              position=position_jitter(0.2)) +
  labs(x = "round", y = "games played") +
  theme_bw()
mosaicplot(table(drafted$round, drafted$season))
```

- Coefficient for 'overall_pick_number' is small, but we can't use these statistics because model does not meet conditions for inference.

- The jitter plot shows that some players drafted in the seventh round played about as many games as those drafted in earlier rounds.


#### Position
```{r}
drafted %>% 
  ggplot(aes(x = factor(position), y = gp)) +
  geom_jitter(alpha = 0.4, aes(color = draft_year),
              position=position_jitter(0.2)) +
  labs(x = "position", y = "games played", color = "draft year") +
  theme_bw()
drafted %>% 
  ggplot(aes(x = factor(position), y = pts)) +
  geom_jitter(alpha = 0.4, aes(color = round),
              position=position_jitter(0.2)) +
  labs(x = "position", y = "points", color = "round") +
  theme_bw()
drafted %>% 
  ggplot(aes(x = factor(position), y = ppg)) +
  geom_jitter(alpha = 0.4, aes(color = round),
              position=position_jitter(0.2)) +
  labs(x = "position", y = "points per game", color = "round") +
  theme_bw()
```

Defensemen have less points and lower points per game than forwards. Players from different draft rounds are intermixed. 

#### ppg in 2019-2020
```{r}
# players with stats from both 2019-2020 and 2021-2022
summed_both <- recent %>% 
  filter(season %in% c("2019-2020", "2021-2022"))
summed_both <- summed_both %>% 
  group_by(name, season) %>% 
  summarize_at(vars(gp:pm, ppg), "sum")
summed_both <- summed_both %>% 
  group_by(name) %>% 
  mutate(count_seasons = n()) %>% 
  filter(count_seasons == 2)

# density plot
summed_both %>%
  ggplot(aes(x = ppg)) +
  geom_density(aes(fill = season), alpha = 0.3) +
  geom_rug(alpha = 0.5) +
  theme_bw()
```

Points per game was slightly higher in the 2019-2020 season. 

## Questions:

1. How do we tell if someone is a first liner?
1. How do we know if it is a tournament or league (since they are the same variable)?
1. Should we be looking at pts or g?

