---
title: "Does Not Playing Hockey Make You Worse At Hockey?"
subtitle: "7/21 Progress Check-In"
author: "Jackie Jovanovic & Michele Sezgin"
output: 
  html_document:
    code_folding: hide
date: 'July 21st, 2022'
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = TRUE)
```

## Goals:
1. get multiple linear regression working
1. establish variables associated with treatment & response and graph to show potential confounders
1. play with transformations
1. look at regression coefficients

#### TAKE OUT BEFORE MEETING

## Introduction
purpose:

WRITE

```{r load-packages, message = FALSE, warning = FALSE}
# load the packages
library(tidyverse)
library(patchwork)
library(lubridate)
library(ggridges)
library(RColorBrewer)
```


```{r data-j, message = FALSE, warning = FALSE}
# load the data
library(readr)
sams_ohl_data_request <- read_csv("C:/Users/jacki/Desktop/REU/hockey/sams_ohl_data_request.csv")
View(sams_ohl_data_request)

# filter years
recent <- sams_ohl_data_request %>% 
  filter(season %in% c("2019-2020", "2020-2021", "2021-2022"))

# add points per game columns
recent <- recent %>% 
  mutate(ppg = g/gp)

# make dataframe of drafted players
drafted <- recent %>% 
  filter(!is.na(draft_year)) %>% 
  mutate(got_drafted = case_when(!is.na(draft_year) ~ 'Yes',
                                 TRUE ~ 'No'))

# only the 2020-2021 season
drafted2021 <- drafted %>% 
  filter(season == "2020-2021")
```

```{r data-m, message = FALSE, warning = FALSE}
ohl <- read_csv("../sams_ohl_data_request.csv")
ohl_covid <- ohl %>%
  filter(season %in% c("2019-2020","2020-2021", "2021-2022"))
```


```{r}
## Add treatment var:
# Filtering only for players who played more than 10 games (should we combine number of games played across leagues?)
ohl_20_21 <- ohl %>%
  filter(season == "2020-2021", gp > 10)
# Split players up into treatment vs non-treatment
ohl_treatment <- ohl %>%
  filter(season %in% c("2019-2020", "2021-2022"), league == "OHL") %>%
  group_by(player_id, season) %>%
  mutate(ppg = sum(pts)/sum(gp),
         treatment = ifelse(player_id %in% ohl_20_21$player_id, "Played", "Didn't play")
  ) %>%
  ungroup()
plyr_quality <- ohl %>%
  filter(season == "2019-2020", league == "OHL") %>%
  group_by(player_id) %>%
  mutate(plyr_quality = sum(pts)/sum(gp)) %>%
  filter(duplicated(player_id) == FALSE) %>%
  ungroup() %>%
  select(player_id, plyr_quality)
# View(plyr_quality %>%
#   filter(duplicated(plyr_quality$player_id) == TRUE))
# ## need to deal with duplicate players/rows
# View(ohl_filtered %>%
#   filter(duplicated(ohl_filtered$player_id) == TRUE))
# We need to check that each player played in the OHL in 2019-2020 and 2021-2022
ohl_szn <- ohl %>%
  filter(season %in% c("2019-2020", "2021-2022"), league == "OHL") %>%
  group_by(player_id) %>%
  mutate(played_2019 = season == "2019-2020",
         played_2021 = season == "2021-2022",
         played_both_szn = sum(played_2019) & sum(played_2021)) %>%
  ungroup() %>%
  filter(played_both_szn == TRUE)
ohl_pm <- ohl_szn %>%
  group_by(team_name, season) %>%
  arrange(team_name, season, desc(pm)) %>%
  mutate(pm_rank = 1:n(),
         pm_relative = pm - mean(pm)/mean(pm)
  ) %>%
  ungroup()
# add treatment variable through join
ohl_trt <- ohl_pm %>%
  left_join(select(ohl_treatment, player_id, season, team_name, treatment), by = c("player_id", "team_name", "season")) %>%
  mutate(ppg = pts/gp) 
# add age variable
ohl_age <- ohl_trt %>%
  group_by(player_id, season) %>%
  mutate(year = strsplit(season, "-")[[1]][1]) %>%
  mutate(age = trunc((dob %--% as.Date(paste0(year, "-9-15"))) / years(1)),
         age_continuous = (dob %--% as.Date(paste0(year, "-09-15"))) / years(1)
  ) %>%
  ungroup() #%>%
  #filter(age %in% 16:22)
    
# add player quality
ohl_qlty <- ohl_age %>%
  left_join(plyr_quality, by = "player_id") %>%
  group_by(player_id, season) %>%
  mutate(ppg_total = sum(pts)/sum(gp),
         gp_total = sum(gp),
         pts_total = sum(pts)) %>%
  arrange(player_id, season, -plyr_quality) %>%
  filter(duplicated(player_id) == FALSE) %>%
  ungroup()#%>%
  # filter(ppg != 0) %>%
  # mutate(log_ppg = log(ppg))
# ohl_qlty <- ohl_age %>%
#   left_join(plyr_quality, by = "player_id") %>%
#   group_by(player_id, season) %>%
#   summarize_at(vars(gp, pts), "sum") %>%
#   group_by(player_id) %>% 
#   mutate(ppg_total = pts/gp, count_seasons = n()) %>% 
#   filter(count_seasons == 2) %>%
#   ungroup()
# were they drafted
ohl_filtered <- ohl_qlty %>%
  mutate(drafted = !is.na(draft_year)) %>%
  select(season, player_id, treatment, first_name, last_name, position, plyr_quality, age, gp_total, pts_total, ppg_total, drafted, draft_year, round, overall_pick_num, age_continuous)
## Trying to combine players who played for two teams into one
# combined <- ohl_filtered %>%
  
ohl_treatment_21_22 <- ohl_filtered %>%
  filter(season == "2021-2022")
```

## EDA Treatment

```{r}
# drafted vs treatment
mosaicplot(table("Drafted" = ohl_filtered$drafted,
                 "Treatment" = ohl_filtered$treatment),
           main = "",
           shade = FALSE)
```

```{r}
ggplot(ohl_filtered, aes(x = gp_total, y = plyr_quality, color = position)) +
  geom_point(alpha = .5)
```

```{r}
# drafted vs treatment
mosaicplot(table("season" = ohl_filtered$season,
                 "Age" = ohl_filtered$age),
           main = "",
           shade = FALSE)
```

```{r}
#continuous age vs ppg
ggplot(ohl_filtered, aes(x = age_continuous, y = ppg_total)) +
  geom_point(alpha = .5)
ohl_filtered %>%
  #filter(age == 18) %>%
ggplot(aes(x = age_continuous, y = ppg_total)) +
  geom_point(alpha = .5) +
  facet_grid(rows = vars(age))
```

## Refit Multiple Linear Regression

```{r code-chunk-label}
mlr_interaction <- lm(ppg_total ~ position*plyr_quality + treatment*drafted + gp_total + plyr_quality*gp_total + plyr_quality*drafted + age_continuous*plyr_quality + season, data = ohl_filtered)
summary(mlr_interaction)
plot(mlr_interaction)
```

```{r}
#played vs didn't play
num_played <- nrow(filter(ohl_filtered, treatment == "Played"))
num_didnt <- nrow(filter(ohl_filtered, treatment == "Didn't play"))
print(paste0("proportion who played is ", round(num_played/nrow(ohl_filtered), 3)))
```

```{r}
# rank within team
rank <- ohl %>%
  filter(league == "OHL", season %in% c("2019-2020", "2021-2022")) %>%
  group_by(team_name, season) %>%
  arrange(team_name, season, desc(pm)) %>%
  mutate(rank = 1:n()) %>%
  ungroup()
relative_pm <- ohl %>%
  filter(league == "OHL", season %in% c("2019-2020", "2021-2022")) %>%
  group_by(team_name, season) %>%
  mutate(rel_pm = pm - mean(pm)/mean(pm)) %>%
  ungroup()
```

```{r}
# glm gamma distribution
set.seed(2004)
ohl_filtered <- ohl_filtered %>%
  mutate(is_train = sample(rep(0:1, length.out = nrow(ohl_filtered))))
#ohl_glm <- glm(ppg_total ~ position + plyr_quality + treatment + drafted + gp_total + age_continuous + season, 
 #                        data = filter(ohl_filtered, is_train == 1), 
  #                       family = Gamma)
```
note: the gamma distribution line is currently commented out because it was giving errors - UPDATE

## Possible Confounding Variables
### Draft
#### Draft Pick Number with Games Played

```{r}
# draft pick # on x-axis, gp on y-axis
drafted %>% 
  ggplot(aes(x = overall_pick_num, y = gp)) +
  geom_point(alpha = 0.5) +
  geom_smooth() +
  labs(x = "draft pick number", y = "games played") +
  theme_bw()
```

```{r}
drafted %>% 
  ggplot(aes(x = gp)) +
  geom_density(aes(color = factor(round)), alpha = 0.3) +
  labs(x = "games played", color = "round") +
  theme_bw()
# still needs work -> thicken line
```

### Goals and Assists

```{r}
goals <- drafted %>% 
  #filter(round %in% c(1, 4, 7)) %>% 
  ggplot(aes(x = g)) +
  geom_density(aes(color = factor(round)), size = .5) +
  labs(x = "goals", color = "round") +
  theme_bw() +
  theme(legend.position = "none")

assists <- drafted %>% 
  #filter(round %in% c(1, 4, 7)) %>% 
  ggplot(aes(x = a)) +
  geom_density(aes(color = factor(round)), size = .5) +
  labs(x = "assists", color = "round") +
  theme_bw()

# shown side-by-side
goals + assists
```

### Plus-Minus

```{r}
teams <- drafted2021 %>% 
  group_by(team_name) %>% 
  mutate(count_players = n()) %>% 
  filter(count_players >= 2)
```

## Questions/Concerns

