---
title: "Does Not Playing Hockey Make You Worse At Hockey?"
subtitle: "7/25 Progress Check-In"
author: "Jackie Jovanovic & Michele Sezgin"
output: 
  html_document:
    code_folding: hide
date: 'July 25th, 2022'
---

<link rel="preconnect" href="https://fonts.googleapis.com">
<link rel="preconnect" href="https://fonts.gstatic.com" crossorigin>
<link href="https://fonts.googleapis.com/css2?family=Open+Sans:wght@300&family=Source+Sans+Pro&display=swap" rel="stylesheet">
<style>

body{
font-family: 'Source Sans Pro', sans-serif;
}

</style>

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = TRUE)
```

## Introduction

Project Motivation: Investigate the impact of pandemic restrictions on hockey player development.

In 2020-2021, many hockey leagues had either a shortened season or no season at all due to COVID-19. When a league, such as the OHL, was shutdown, players had to find different leagues and/or tournaments to participate in. Some players were unable to practice with a team for that season. How did this disruption in training influence a player's development? To answer this question, we will examine data from the 2019-2020, 2020-2021, and 2021-2022 seasons for junior leagues.

```{r load-packages, message = FALSE, warning = FALSE}
library(tidyverse)
library(lubridate)
library(MatchIt)
library(rpart)
library(rpart.plot)
library(caret)
library(vip)
library(RColorBrewer)
library(viridis)
```

```{r load-data, message = FALSE, warning = FALSE}
library(readr)
sams_ohl_data_request <- read_csv("sams_ohl_data_request.csv")
View(sams_ohl_data_request)

# filter years
recent <- sams_ohl_data_request %>% 
  filter(season %in% c("2019-2020", "2020-2021", "2021-2022"))
# add points per game columns
recent <- recent %>% 
  mutate(ppg = g/gp)
# add drafted y/n column 
recent <- recent %>% 
  mutate(got_drafted = case_when(!is.na(draft_year) ~ 'Yes',
                                 TRUE ~ 'No'))
# yes is only plays drafted before COVID
recent2 <- recent %>% 
  mutate(got_drafted = case_when(!is.na(draft_year) & draft_year < 2020
                                 ~ 'Yes',
                                 TRUE ~ 'No'))

ranks <- recent2 %>% 
  group_by(team_name, season) %>% 
  arrange(team_name, season, desc(pm)) %>% 
  mutate(pm_rank = 1:n(),
         pm_mean = mean(pm),
         pm_relative = pm - mean(pm)) %>% 
  ungroup()

# filter out NA values
recent3 <- recent2 %>% 
  filter(!is.na(ppg), !is.na(g), !is.na(pts), !is.na(pm), !is.na(gp),
         !is.na(a), !is.na(pim))
ranks2 <- ranks %>% 
  filter(!is.na(ppg), !is.na(g), !is.na(pts), !is.na(pm), !is.na(gp),
         !is.na(a), !is.na(pim))
```

```{r treatment, message = FALSE, warning = FALSE}
# copied from Michele's midpoint code

# Filtering only for players who played more than 10 games 
# (should we combine number of games played across leagues?)
ohl_20_21 <- recent %>%
  filter(season == "2020-2021", gp > 10)
# Split players up into treatment vs non-treatment
ohl_treatment <- recent %>%
  filter(season %in% c("2019-2020", "2021-2022"), league == "OHL") %>%
  group_by(player_id, season) %>%
  mutate(ppg = sum(pts)/sum(gp),
         treatment = ifelse(player_id %in% ohl_20_21$player_id, "Played", "Didn't play")
  ) %>%
  ungroup()
plyr_quality <- recent %>%
  filter(season == "2019-2020", league == "OHL") %>%
  group_by(player_id) %>%
  mutate(plyr_quality = sum(pts)/sum(gp)) %>%
  filter(duplicated(player_id) == FALSE) %>%
  ungroup() %>%
  select(player_id, plyr_quality)
ohl_pm <- recent %>%
  filter(season %in% c("2019-2020", "2021-2022"), league == "OHL") %>%
  group_by(team_name, season) %>%
  arrange(team_name, season, desc(pm)) %>%
  mutate(pm_rank = 1:n(),
         pm_mean = mean(pm),
         pm_relative = pm - mean(pm)
  ) %>%
  ungroup()
# We need to check that each player played in the OHL in 2019-2020 and 2021-2022
ohl_szn <- ohl_pm %>%
  group_by(player_id) %>%
  mutate(played_2019 = season == "2019-2020",
         played_2021 = season == "2021-2022",
         played_both_szn = sum(played_2019) & sum(played_2021)) %>%
  ungroup() %>%
  filter(played_both_szn == TRUE)
# add treatment variable through join
ohl_trt <- ohl_szn %>%
  left_join(select(ohl_treatment, player_id, season, team_name, treatment), by = c("player_id", "team_name", "season")) %>%
  mutate(ppg = pts/gp) 
# add age variable
ohl_age <- ohl_trt %>%
  group_by(player_id, season) %>%
  mutate(year = strsplit(season, "-")[[1]][1]) %>%
  mutate(age = trunc((dob %--% as.Date(paste0(year, "-9-15"))) / years(1)),
         age_continuous = (dob %--% as.Date(paste0(year, "-09-15"))) / years(1)
  ) %>%
  ungroup()
# add player quality
ohl_qlty <- ohl_age %>%
  left_join(plyr_quality, by = "player_id") %>%
  group_by(player_id, season) %>%
  mutate(ppg_total = sum(pts)/sum(gp),
         gp_total = sum(gp),
         pts_total = sum(pts)) %>%
  arrange(player_id, season, -plyr_quality) %>%
  filter(duplicated(player_id) == FALSE) %>%
  ungroup()
# were they drafted
ohl_filtered <- ohl_qlty %>%
  mutate(drafted = ifelse(!is.na(draft_year) == TRUE, "Yes", "No")) %>%
  select(season, player_id, treatment, first_name, last_name, position, plyr_quality, age, gp_total, pts_total, ppg_total, drafted, draft_year, round, overall_pick_num, age_continuous, pm_relative, pm_rank, pm)
```

## Matching

```{r matching, message = FALSE, warning = FALSE, fig.align = 'center'}
# make a binary treatment variable
filtered <- ohl_filtered %>% 
  mutate(num_treatment = case_when(treatment == "Didn't play" ~ 0,
                                   TRUE ~ 1))

# first matching attempt
match1 <-
  matchit(num_treatment ~ position + gp_total + pts_total + ppg_total + drafted +
            pm_relative + pm_rank + pm,
          data = filtered, method = "nearest",
          distance = "mahalanobis",
          replace = FALSE, ratio = 2)
summary(match1)
plot(summary(match1)) # love plot of the covariates

gam_propensity_match1 <- 
  matchit(num_treatment ~ position + s(gp_total) + s(pts_total) + 
            s(ppg_total) + drafted + s(pm_relative) + s(pm_rank) + s(pm),
          data = filtered, method = "nearest",
          distance = "gam",
          replace = FALSE, ratio = 2)
summary(gam_propensity_match1)
plot(gam_propensity_match1, type = "jitter", interactive = FALSE)
```

Graph interpretation? Is the data matched up enough?

## Exploring Different Measurements of Player Quality

```{r playerq, message = FALSE, warning = FALSE, fig.align = 'center'}
# find first-liners
#ranks %>% 
 # filter(position == "F" & pm_rank == 1:6)
#ranks %>% 
 # filter(position == "D" & pm_rank == 1:2) %>% View()
```

### Plus-Minus {.tabset}

#### Tournaments
```{r, message = FALSE, warning = FALSE, fig.align = 'center'}
ranks2 %>% 
  filter(league %in% c("International-Jr", "WJC-18", "WJC-20")) %>% 
  ggplot(aes(x = player_id, y = pm_relative)) +
  geom_point(aes(color = league), alpha = 0.8) +
  geom_hline(aes(yintercept = mean(pm_relative)), color = "darkgoldenrod3") +
  geom_hline(aes(yintercept = 0), color = "goldenrod1") +
  labs(title = "Tournaments", y = "Relative Plus-Minus") +
  scale_color_brewer(palette = "Blues") +
  theme_bw() +
  theme(axis.text.x = element_blank(), axis.ticks.x = element_blank(),
        axis.title.x = element_blank()) +
  theme(legend.position = "bottom") +
  facet_wrap(~season)
```

#### AHL
```{r ahl, message = FALSE, warning = FALSE, fig.align = 'center'}
ranks2 %>% 
  filter(league %in% c("AHL")) %>% 
  ggplot(aes(x = player_id, y = pm_relative)) +
  geom_point(alpha = 0.5) +
  geom_hline(aes(yintercept = mean(pm_relative)), color = "darkgoldenrod3") +
  geom_hline(aes(yintercept = 0), color = "royalblue3", alpha = 0.5) +
  labs(title = "AHL", y = "Relative Plus-Minus", 
       caption = "Mean plus-minus is 0.") +
  theme_bw() +
  theme(axis.text.x = element_blank(), axis.ticks.x = element_blank(),
        axis.title.x = element_blank()) +
  theme(legend.position = "bottom") +
  facet_wrap(~season)
```

The AHL does not have any games from the 2019-2020 season.

#### OHL
```{r ohl, message = FALSE, warning = FALSE, fig.align = 'center'}
ranks2 %>% 
  filter(league %in% c("OHL")) %>% 
  ggplot(aes(x = player_id, y = pm_relative)) +
  geom_point(alpha = 0.5) +
  geom_hline(aes(yintercept = mean(pm_relative)), color = "darkgoldenrod3") +
  geom_hline(aes(yintercept = 0), color = "royalblue3", alpha = 0.5) +
  labs(title = "OHL", y = "Relative Plus-Minus", 
       caption = "Mean plus-minus is 0.2016.") +
  theme_bw() +
  theme(axis.text.x = element_blank(), axis.ticks.x = element_blank(),
        axis.title.x = element_blank()) +
  theme(legend.position = "bottom") +
  facet_wrap(~season)
```

### {-}

## Differences Between Leagues
```{r}
# summary statistics for each league
leagues1920 <- recent %>% 
  filter(season == "2019-2020") %>% 
  group_by(league) %>% 
  summarize(count = n(), gp = mean(gp), pts = mean(pts), pm = mean(pm),
            ppg = mean(ppg)) 
leagues2021 <- recent %>% 
  filter(season == "2020-2021") %>% 
  group_by(league) %>% 
  summarize(count = n(), gp = mean(gp), pts = mean(pts), pm = mean(pm),
            ppg = mean(ppg))
leagues2122 <- recent %>% 
  filter(season == "2021-2022") %>% 
  group_by(league) %>% 
  summarize(count = n(), gp = mean(gp), pts = mean(pts), pm = mean(pm),
            ppg = mean(ppg))
```

### Points vs. Games Played {.tabset}

#### 2019-2020
```{r, message=FALSE, warning=FALSE, fig.align='center'}
leagues1920 %>% 
  ggplot(aes(x = gp, y = pts)) +
  geom_point(aes(color = factor(count))) +
  scale_color_viridis(discrete = TRUE) +
  labs(title = "2019-2020 Season", 
       x = "Games Played", y = "Points", color = "Game Count") +
  theme_bw() 
```

#### 2020-2021
```{r, message=FALSE, warning=FALSE, fig.align='center'}
leagues2021 %>% 
  ggplot(aes(x = gp, y = pts)) +
  geom_point(aes(color = factor(count))) +
  scale_color_viridis(discrete = TRUE) +
  labs(title = "2020-2021 Season", 
       x = "Games Played", y = "Points", color = "Game Count") +
  theme_bw() 
```

#### 2021-2022
```{r, message=FALSE, warning=FALSE, fig.align='center'}
leagues2122 %>% 
  ggplot(aes(x = gp, y = pts)) +
  geom_point(aes(color = factor(count))) +
  scale_color_viridis(discrete = TRUE) +
  labs(title = "2021-2022 Season", 
       x = "Games Played", y = "Points", color = "Game Count") +
  theme_bw() 
```

### {-}

### Histograms of Counts {.tabset}

#### 2019-2020
```{r, message=FALSE, warning=FALSE, fig.align='center'}
leagues1920 %>% 
  filter(league != "OHL") %>% 
  ggplot(aes(x = count)) +
  geom_histogram(color = "white") + 
  labs(x = "Leagues with Observed Game Count", y = "Count",
       title = "2019-2020 Season",
       caption = "36 leagues have one recorded game.
       OHL has been omitted, they have a count of 510.") +
  annotate("text", x = 32.5, y = 31, label = "97 Leagues",
           col = "royalblue3") +
  theme_bw()
```


#### 2020-2021
```{r, message=FALSE, warning=FALSE, fig.align='center'}
leagues2021 %>% 
  ggplot(aes(x = count)) +
  geom_histogram(color = "white") + 
  labs(x = "Leagues with Observed Game Count", y = "Count",
       title = "2020-2021 Season",
       caption = "31 leagues have one recorded game.") +
  annotate("text", x = 21.25, y = 26, label = "71 Leagues",
           col = "royalblue3") +
  theme_bw()
```


#### 2021-2022
```{r, message=FALSE, warning=FALSE, fig.align='center'}
leagues2122 %>% 
  filter(league != "OHL") %>% 
  ggplot(aes(x = count)) +
  geom_histogram(color = "white") + 
  labs(x = "Leagues with Observed Game Count", y = "Count",
       title = "2021-2022 Season",
       caption = "27 leagues have one recorded game.
       OHL has been omitted, they have a count of 534.") +
  annotate("text", x = 22.5, y = 21, label = "49 Leagues",
           col = "royalblue3") +
  theme_bw()
```


### {-}


## Regression Trees

### Tree 1 {.tabset}

**formula: ppg ~ gp + g + pts + pm**

#### Initial
```{r message = FALSE, warning = FALSE, fig.align = 'center'}
#tree 1
tree1 <- rpart(formula = ppg ~ gp + g + pts + pm,
              data = recent3, method = "anova")
#tree1 <- rpart(formula = ppg ~ gp + g + pts + pm + ppg + factor(got_drafted) +
    #            factor(position),
     #         data = recent2, method = "anova") # do the factors change?
rpart.plot(tree1)
```

#### Trained
```{r message = FALSE, warning = FALSE, fig.align = 'center'}
# train the data
caret_tree1 <- train(ppg ~ gp + g + pts + pm,
                     data = recent3, method = "rpart",
                     trControl = trainControl(method = "cv", number = 10),
                     tuneLength = 20)
# what do the bottom two lines do?
rpart.plot(caret_tree1$finalModel)
# why is it only gp and g?
```

#### Variable Importance
```{r message = FALSE, warning = FALSE, fig.align = 'center'}
vip(caret_tree1, geom = "point") + theme_bw()
```

### {-}

Questions embedded in code. 

### Tree 2 {.tabset}

**formula: pm ~ gp + g + a + pim + got_drafted**

#### Initial
```{r message = FALSE, warning = FALSE, fig.align = 'center'}
tree2 <- rpart(formula = pm ~ gp + g + a + pim + got_drafted,
               data = recent3, method = "anova")
rpart.plot(tree2)
```

#### Trained
```{r message = FALSE, warning = FALSE, fig.align = 'center'}
caret_tree2 <- train(pm ~ gp + g + a + pim + got_drafted,
                     data = recent3, method = "rpart",
                     trControl = trainControl(method = "cv", number = 10),
                     tuneLength = 20)
rpart.plot(caret_tree2$finalModel)
```

#### Variable Importance
```{r message = FALSE, warning = FALSE, fig.align = 'center'}
vip(caret_tree2, geom = "point") + theme_bw()
```

### {-}

### Tree 3 {.tabset}

**formula: ppg ~ gp + g + pts + pm**

#### Initial
```{r message = FALSE, warning = FALSE, fig.align = 'center'}
tree3 <- rpart(formula = ppg ~ gp + g + pts + pm,
               data = ranks2, method = "anova")
rpart.plot(tree3)
```

#### Trained
```{r message = FALSE, warning = FALSE, fig.align = 'center'}
caret_tree3 <- train(ppg ~ gp + g + pts + pm,
                     data = ranks2, method = "rpart",
                     trControl = trainControl(method = "cv", number = 10),
                     tuneLength = 20)
rpart.plot(caret_tree3$finalModel)
```

#### Variable Importance
```{r message = FALSE, warning = FALSE, fig.align = 'center'}
vip(caret_tree3, geom = "point") + theme_bw()
```

### {-}

### Tree 4 {.tabset}

**formula: pm ~ gp + g + a + pim + got_drafted**

#### Initial
```{r message = FALSE, warning = FALSE, fig.align = 'center'}
tree4 <- rpart(formula = pm ~ gp + g + a + pim + got_drafted,
               data = ranks2, method = "anova")
rpart.plot(tree4)
```

#### Trained
```{r message = FALSE, warning = FALSE, fig.align = 'center'}
caret_tree4 <- train(pm ~ gp + g + a + pim + got_drafted,
                     data = ranks2, method = "rpart",
                     trControl = trainControl(method = "cv", number = 10),
                     tuneLength = 20)
rpart.plot(caret_tree4$finalModel)
```

#### Variable Importance
```{r message = FALSE, warning = FALSE, fig.align = 'center'}
vip(caret_tree4, geom = "point") + theme_bw()
```

### {-}

### Tree 5 {.tabset}

**formula: ppg_total ~ treatment + position + age_continuous + gp_total + pts_total + drafted + pm_relative + pm_rank + pm**

#### Initial
```{r message = FALSE, warning = FALSE, fig.align = 'center'}
tree5 <- rpart(formula = ppg_total ~ treatment + position + age_continuous +
                 gp_total + pts_total + drafted + pm_relative + pm_rank +
                 pm, 
               data = filtered, method = "anova")
rpart.plot(tree5)
```

#### Trained
```{r message = FALSE, warning = FALSE, fig.align = 'center'}
caret_tree5 <- train(ppg_total ~ treatment + position + age_continuous +
                       gp_total + pts_total + drafted + pm_relative + 
                       pm_rank + pm,
                     data = filtered, method = "rpart",
                     trControl = trainControl(method = "cv", number = 10),
                     tuneLength = 20)
rpart.plot(caret_tree5$finalModel)
```

#### Variable Importance
```{r message = FALSE, warning = FALSE, fig.align = 'center'}
vip(caret_tree5, geom = "point") + theme_bw()
```

### {-}

## Questions

1. What do I put for x.train and y.train in BART?
1. Questions embedded in Tree 1 code.
1. How can I make a plot interactive? (scroll over bar to make league name visible)

