---
title: "Does Not Playing Hockey Make You Worse At Hockey?"
subtitle: "Midpoint Draft"
author: "Jackie Jovanovic & Michele Sezgin"
output: 
  html_document:
    code_folding: hide
date: 'July 22nd, 2022'
---

<link rel="preconnect" href="https://fonts.googleapis.com">
<link rel="preconnect" href="https://fonts.gstatic.com" crossorigin>
<link href="https://fonts.googleapis.com/css2?family=Open+Sans:wght@300&family=Source+Sans+Pro&display=swap" rel="stylesheet">
<style>

body{
font-family: 'Source Sans Pro', sans-serif;
}

</style>

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = TRUE)
```

# Introduction 

Project Motivation: Investigate the impact of pandemic restrictions on hockey player development.

*   2020-2021 many leagues had a shortened season or no season
*   Players had to find new leagues and/or tournament
*   Some players did not play that season

```{r load-packages, include = FALSE}
# load the packages
library(tidyverse)
library(patchwork)
library(lubridate)
library(ggridges)
library(RColorBrewer)
library(kableExtra)
```

```{r data-jackie, message = FALSE, warning = FALSE}
# load the data
library(readr)
sams_ohl_data_request <- read_csv("C:/Users/jacki/Desktop/REU/hockey/sams_ohl_data_request.csv")
View(sams_ohl_data_request)

# filter years
recent <- sams_ohl_data_request %>% 
  filter(season %in% c("2019-2020", "2020-2021", "2021-2022"))

# add points per game columns
recent <- recent %>% 
  mutate(ppg = g/gp)

# add drafted y/n column 
recent <- recent %>% 
  mutate(got_drafted = case_when(!is.na(draft_year) ~ 'Yes',
                                 TRUE ~ 'No'))

# make dataframe of drafted players
drafted <- recent %>% 
  filter(!is.na(draft_year)) %>% 
  filter(draft_year < 2020)
```

```{r data-michele, message = FALSE, warning = FALSE}
ohl <- read_csv("../sams_ohl_data_request.csv")
ohl_covid <- ohl %>%
  filter(season %in% c("2019-2020","2020-2021", "2021-2022"))

## Add treatment var:
# Filtering only for players who played more than 10 games (should we combine number of games played across leagues?)
ohl_20_21 <- ohl %>%
  filter(season == "2020-2021", gp > 10)
# Split players up into treatment vs non-treatment
ohl_treatment <- ohl %>%
  filter(season %in% c("2019-2020", "2021-2022"), league == "OHL") %>%
  group_by(player_id, season) %>%
  mutate(ppg = sum(pts)/sum(gp),
         treatment = ifelse(player_id %in% ohl_20_21$player_id, "Played", "Didn't play")
  ) %>%
  ungroup()
plyr_quality <- ohl %>%
  filter(season == "2019-2020", league == "OHL") %>%
  group_by(player_id) %>%
  mutate(plyr_quality = sum(pts)/sum(gp)) %>%
  filter(duplicated(player_id) == FALSE) %>%
  ungroup() %>%
  select(player_id, plyr_quality)
# View(plyr_quality %>%
#   filter(duplicated(plyr_quality$player_id) == TRUE))
# ## need to deal with duplicate players/rows
# View(ohl_filtered %>%
#   filter(duplicated(ohl_filtered$player_id) == TRUE))
ohl_pm <- ohl %>%
  filter(season %in% c("2019-2020", "2021-2022"), league == "OHL") %>%
  group_by(team_name, season) %>%
  arrange(team_name, season, desc(pm)) %>%
  mutate(pm_rank = 1:n(),
         pm_relative = pm - mean(pm)/mean(pm)
  ) %>%
  ungroup()
# We need to check that each player played in the OHL in 2019-2020 and 2021-2022
ohl_szn <- ohl_pm %>%
  group_by(player_id) %>%
  mutate(played_2019 = season == "2019-2020",
         played_2021 = season == "2021-2022",
         played_both_szn = sum(played_2019) & sum(played_2021)) %>%
  ungroup() %>%
  filter(played_both_szn == TRUE)
# add treatment variable through join
ohl_trt <- ohl_szn %>%
  left_join(select(ohl_treatment, player_id, season, team_name, treatment), by = c("player_id", "team_name", "season")) %>%
  mutate(ppg = pts/gp) 
# add age variable
ohl_age <- ohl_trt %>%
  group_by(player_id, season) %>%
  mutate(year = strsplit(season, "-")[[1]][1]) %>%
  mutate(age = trunc((dob %--% as.Date(paste0(year, "-9-15"))) / years(1)),
         age_continuous = (dob %--% as.Date(paste0(year, "-09-15"))) / years(1)
  ) %>%
  ungroup() #%>%
  #filter(age %in% 16:22)
    
# add player quality
ohl_qlty <- ohl_age %>%
  left_join(plyr_quality, by = "player_id") %>%
  group_by(player_id, season) %>%
  mutate(ppg_total = sum(pts)/sum(gp),
         gp_total = sum(gp),
         pts_total = sum(pts)) %>%
  arrange(player_id, season, -plyr_quality) %>%
  filter(duplicated(player_id) == FALSE) %>%
  ungroup()#%>%
  # filter(ppg != 0) %>%
  # mutate(log_ppg = log(ppg))
# ohl_qlty <- ohl_age %>%
#   left_join(plyr_quality, by = "player_id") %>%
#   group_by(player_id, season) %>%
#   summarize_at(vars(gp, pts), "sum") %>%
#   group_by(player_id) %>% 
#   mutate(ppg_total = pts/gp, count_seasons = n()) %>% 
#   filter(count_seasons == 2) %>%
#   ungroup()
# were they drafted
ohl_filtered <- ohl_qlty %>%
  mutate(drafted = !is.na(draft_year)) %>%
  select(season, player_id, treatment, first_name, last_name, position, plyr_quality, age, gp_total, pts_total, ppg_total, drafted, draft_year, round, overall_pick_num, age_continuous, pm_relative, pm_rank, pm)
## Trying to combine players who played for two teams into one
# combined <- ohl_filtered %>%
  
ohl_treatment_21_22 <- ohl_filtered %>%
  filter(season == "2021-2022")
```

# Data

We are looking at data from the 2019-2020, 2020-2021, and 2021-2022 seasons for junior leagues across Canada and Europe.

<span style="color: darkblue;">
**Raw Data:**
</span>

```{r, message = FALSE, warning = FALSE}
recent %>% 
  dplyr::select(name, team_name, season, league, position, 
                gp, g, a, pts, pm) %>%
  head(5) %>% 
  knitr::kable()%>% 
  kable_styling("striped")
```

<span style="color: darkblue;">
**Player Example:**
</span>

```{r, message = FALSE, warning = FALSE}
recent %>% 
  dplyr::select(name, team_name, season, league, position, 
                gp, g, a, pts, pm) %>%
  filter(name == "Brett Harrison") %>%
  arrange(season) %>% 
  knitr::kable() %>% 
  kable_styling("striped")
```

# Exploratory Data Analysis

<span style="color: darkblue;">
**Points Per Game and Games Played**
</span>

```{r, fig.align = 'center', fig.width = 10}
ppg_ohl <- recent %>% 
  filter(league == "OHL") %>%
  filter(ppg > 0) %>% 
  ggplot(aes(x = gp, y = ppg)) +
  geom_point(aes(color = got_drafted), alpha = 0.3) +
  labs(title = "OHL", x = "Games Played", y = "Points Per Game",
       color = "Drafted Before \n 2020-2021 Season") +
  theme_bw() +
  theme(legend.position = "none")
ppg_all <- recent %>% 
  filter(ppg > 0) %>% 
  ggplot(aes(x = gp, y = ppg)) +
  geom_point(aes(color = got_drafted), alpha = 0.3) +
  labs(title = "All Leagues", x = "Games Played", y = "Points Per Game",
       color = "Drafted Before \n 2020-2021 Season") +
  theme_bw()
ppg_ohl + ppg_all
```


