---
title: "Does Not Playing Hockey Make You Worse At Hockey?"
subtitle: "BART and Matching"
author: "Jackie Jovanovic & Michele Sezgin"
output: 
  html_document:
    code_folding: hide
date: 'July 27th, 2022'
---

<link rel="preconnect" href="https://fonts.googleapis.com">
<link rel="preconnect" href="https://fonts.gstatic.com" crossorigin>
<link href="https://fonts.googleapis.com/css2?family=Open+Sans:wght@300&family=Source+Sans+Pro&display=swap" rel="stylesheet">
<style>

body{
font-family: 'Source Sans Pro', sans-serif;
}

</style>

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = TRUE)
```

## Introduction

Project Motivation: Investigate the impact of pandemic restrictions on hockey player development.

In 2020-2021, many hockey leagues had either a shortened season or no season at all due to COVID-19. When a league, such as the OHL, was shutdown, players had to find different leagues and/or tournaments to participate in. Some players were unable to practice with a team for that season. How did this disruption in training influence a player's development? To answer this question, we will examine data from the 2019-2020, 2020-2021, and 2021-2022 seasons for junior leagues.

```{r load-packages, message = FALSE, warning = FALSE}
library(tidyverse)
library(BART)
library(tidytreatment)
library(dplyr)
library(tidybayes)
library(ggplot2)
library(MatchIt)
library(cobalt)
library(mgcv)
library(lubridate)
```


```{r code-chunk-label}
# Rethink data

ohl <- read_csv("sams_ohl_data_request.csv")

# Create df of players who played during covid w total gp and if they played championship
ohl_treatment_prep <- ohl %>%
  filter(season == "2020-2021") %>%
  group_by(player_id) %>%
  mutate(gp_covid = sum(gp),
         championship = ifelse(league %in% c("WC","WJC-18", "WJC-20"), TRUE, FALSE),
         covid_league = league) %>%
  ungroup() %>%
  dplyr::select(player_id, gp_covid, championship, covid_league) %>%
  filter(duplicated(player_id) == FALSE)

# Compute player quality vars using pm
ohl_pm <- ohl %>%
  filter(season == "2019-2020", league == "OHL") %>%
  group_by(team_name) %>%
  arrange(team_name, desc(pm)) %>%
  mutate(pm_rank = 1:n(),
         pm_relative = pm - mean(pm)
  ) %>%
  ungroup() %>%
  dplyr::select(player_id, team_name, pm_rank, pm_relative)

# We need to check that each player played in the OHL in 2019-2020 and 2021-2022
ohl_szn <- ohl %>%
  filter(season %in% c("2019-2020", "2021-2022"), league == "OHL") %>%
  group_by(player_id) %>%
  mutate(played_2019 = season == "2019-2020",
         played_2021 = season == "2021-2022",
         played_both_szn = sum(played_2019) & sum(played_2021)) %>%
  ungroup() %>%
  filter(played_both_szn == TRUE)
  
ohl_21_22_ppg <- ohl_szn %>%
  filter(season == "2021-2022") %>%
  group_by(player_id) %>%
  filter(duplicated(player_id) == FALSE) %>%
  mutate(ppg_21_22 = sum(pts)/sum(gp),
         gp_21_22 = sum(gp)) %>%
  dplyr::select(player_id, ppg_21_22, gp_21_22)

# Add treatment vars and pm
ohl_treatment <- ohl_szn %>%
  filter(season == "2019-2020") %>%
  left_join(ohl_treatment_prep, by = "player_id") %>%
  left_join(ohl_pm, by = c("player_id", "team_name")) %>%
  replace_na(list(gp_covid = 0, championship = FALSE))

# add age variable
ohl_age <- ohl_treatment %>%
  group_by(player_id) %>%
  mutate(age = trunc((dob %--% as.Date("2020-01-01")) / years(1)),
         age_continuous = (dob %--% as.Date("2020-01-01")) / years(1)
  ) %>%
  ungroup()

# summarizing variables, adding player performance as ppg
ohl_performance <- ohl_age %>%
  group_by(player_id) %>%
  mutate(ppg_19_20 = sum(pts)/sum(gp),
         gp_19_20 = sum(gp),
         pts_19_20 = sum(pts),
         pm_relative_19_20 = mean(pm_relative),
         pm_rank_19_20 = trunc(mean(pm_rank)),
         pm_19_20 = pm,
         treatment = ifelse(gp_covid > 0, "Played", "Didn't Play")
  ) %>%
  arrange(player_id, season) %>%
  filter(duplicated(player_id) == FALSE) %>%
  ungroup()

# were they drafted?
ohl_drafted <- ohl_performance %>%
  mutate(drafted = case_when(draft_year != 2021 ~ TRUE,
                             draft_year == NA ~ FALSE,
                             TRUE ~ FALSE))

# add 2021 ppg
ohl_filtered <- ohl_drafted %>%
  left_join(ohl_21_22_ppg, by = "player_id")

# clean up
ohl_filtered <- ohl_filtered %>%
  dplyr::select(season, player_id, championship, gp_19_20, first_name, last_name, position, age, gp_19_20, pts_19_20, ppg_19_20, drafted, draft_year, round, overall_pick_num, age_continuous, pm_relative_19_20, pm_rank_19_20, pm_19_20, treatment, ppg_21_22, gp_covid, gp_21_22, covid_league)
```


```{r data, message = FALSE, warning = FALSE}
library(readr)

#ohl_filtered <- read_csv("ohl_filtered.csv")

ohl_filtered <- ohl_filtered %>% 
  mutate(got_drafted = case_when(!is.na(draft_year) & draft_year < 2020
                                 ~ 'Yes',
                                 TRUE ~ 'No'))

# Convert treatment to binary indicator:
ohl_filtered <- ohl_filtered %>%
  mutate(treatment_num = ifelse(treatment == "Played", 1, 0))

ohl_update <- ohl_filtered %>% 
  mutate(is_forward = case_when(position == "F" ~ 1, 
                                position == "D" ~ 0)) %>% 
  mutate(is_drafted = case_when(got_drafted == "Yes" ~ 1,
                                got_drafted == "No" ~ 0)) %>% 
  # mutate(played_covid = case_when(treatment == "TRUE" ~ 1,
  #                                treatment == "FALSE" ~ 0)) %>% 
  mutate(s_played = case_when(season == "2021-2022" ~ 1,
                              season == "2019-2020" ~ 0)) %>% 
  as.data.frame()
```

## Using tidytreatment with BART

```{r fit-var-sel-model, message = FALSE, warning = FALSE}
### fit variable selection model
var_select_bart <- wbart(x.train = select(ohl_update, gp_19_20, pts_19_20, 
                                         age_continuous, pm_relative_19_20, 
                                         pm_rank_19_20, pm_19_20, is_forward, 
                                         is_drafted),
                        y.train = pull(ohl_update, ppg_19_20),
                        sparse = TRUE,
                        nskip = 2000,
                        ndpost = 5000)

### variable selection
covar_ranking <- covariate_importance(var_select_bart)
var_select <- covar_ranking %>% 
  filter(avg_inclusion >= quantile(avg_inclusion, 0.5)) %>% 
  pull(variable)

### fit a propensity score model
prop_bart <- pbart(x.train = select(ohl_update, all_of(var_select)),
                   y.train = pull(ohl_update, treatment),
                   nskip = 2000,
                   ndpost = 5000)

### store propensity score in data
ohl_update$prop_score <- prop_bart$prob.train.mean

### fit the treatment effect model
te_model <- wbart(x.train = select(ohl_update, gp_19_20, pts_19_20, 
                                   age_continuous, pm_relative_19_20, 
                                   pm_rank_19_20, pm_19_20, is_forward, 
                                   is_forward, 
                                   is_drafted, treatment_num, prop_score),
                                   # need to include treatment and prop score)
                  y.train = pull(ohl_update, ppg_19_20),
                  nskip = 10000L,
                  ndpost = 200L,
                  keepevery = 100L)

### extract the posterior
posterior_fitted <- fitted_draws(te_model, value = "fit",
                                 include_newdata = FALSE)
posterior_fitted

# function to tidy predicted draws and add random normal noise by default

posterior_pred <- predicted_draws(te_model, include_newdata = FALSE)

### plots using tidybayes
treatment_var_and_c1 <- ohl_update %>% 
  dplyr::select(treatment_num, is_forward) %>% 
  mutate(.row = 1:n(), treatment_num = as.factor(treatment_num))

posterior_fitted %>% 
  left_join(treatment_var_and_c1, by = ".row") %>% 
  ggplot() +
  stat_halfeye(aes(x = treatment, y = fit)) +
  facet_wrap(~is_forward, labeller = 
               as_labeller( function(x) paste("is_forward =",x) ) ) +
  xlab("Treatment (played_covid)") + ylab("Posterior predicted value") +
  theme_bw() + 
  ggtitle("Effect of treatment with 'is_forward' on posterior fitted value")

### calculate the treatment effects
# sample based (using data from fit) conditional treatment effects,
# posterior draws
posterior_treat_eff <- 
  treatment_effects(te_model, treatment = "treatment", 
                    # The dataset here needs to match the BART data EXACTLY
                    # which is really annoying...
                    newdata = select(ohl_update, gp_19_20, pts_19_20, 
                                     age_continuous, pm_relative_19_20, 
                                     pm_rank_19_20, pm_19_20, is_forward, 
                                     is_drafted,
                                     treatment, prop_score))

# Now plot:
posterior_treat_eff %>% 
  ggplot() +
  geom_histogram(aes(x = cte), bins = 30, colour = "white") + 
  theme_bw() + ggtitle("Histogram of treatment effect (all draws)")

```

## Matching

```{r include=FALSE}
# ohl_filtered_m <- read_csv("ohl_filtered.csv")
# 
# ohl_filtered_m <- ohl_filtered_m %>% 
#   mutate(got_drafted = case_when(!is.na(draft_year) & draft_year < 2020
#                                  ~ 'Yes',
#                                  TRUE ~ 'No'))
# ohl_filtered2 <- ohl_filtered_m
# 
# # from bart -> make treatment binary
# ohl_filtered2 <- ohl_filtered2 %>% 
#   mutate(treatment = ifelse(treatment == "Played", 1, 0),
#          treatment = as.integer((treatment)))
```


Includes the most variables:
```{r message = FALSE, warning = FALSE, fig.align='center'}

# ohl_filtered2 <- ohl_filtered2 %>% 
#   mutate(treatment = ifelse(treatment == "Played", 1, 0),
#          treatment = as.integer((treatment)))

gam_propensity_match <- 
  matchit(treatment_num ~ gp_19_20 + position + pts_19_20 + ppg_19_20 + 
            age_continuous + pm_19_20, 
          data = ohl_filtered, method = "nearest",
          distance = "gam",
          replace = FALSE, # do not reuse controls
          ratio = 1)

summary(gam_propensity_match)
plot(gam_propensity_match, type = "jitter", interactive = FALSE)
```

Does not have pts_19_20:
```{r message = FALSE, warning = FALSE, fig.align='center'}
# does not have pts_19_20
gam_propensity_match2 <- 
  matchit(treatment_num ~ gp_19_20 + position + ppg_19_20 + 
            age_continuous + pm_19_20, 
          data = ohl_filtered, method = "nearest",
          distance = "gam",
          replace = FALSE, # do not reuse controls
          ratio = 1)
summary(gam_propensity_match2) # lowest Std. Pair Dist.
plot(gam_propensity_match2, type = "jitter", interactive = FALSE)
plot(summary(gam_propensity_match2))
```

Does not have pts_19_20 or pm_19_20:
```{r message = FALSE, warning = FALSE, fig.align='center'}
# does not have pm_19_20
gam_propensity_match3 <- 
  matchit(treatment_num ~ gp_19_20 + position + ppg_19_20 + 
            age_continuous, 
          data = ohl_filtered, method = "nearest",
          distance = "gam",
          replace = FALSE, # do not reuse controls
          ratio = 1)
summary(gam_propensity_match3)
plot(gam_propensity_match3, type = "jitter", interactive = FALSE)
plot(summary(gam_propensity_match3))
```

```{r message = FALSE, warning = FALSE, fig.align='center'}
# just matched data
gam_matched <- match.data(gam_propensity_match)
table(gam_matched$treatment)
```


```{r message = FALSE, warning = FALSE, fig.align='center'}
bal.plot(gam_propensity_match, var.name = "distance") # look into
bal.plot(gam_propensity_match2, var.name = "distance")
bal.plot(gam_propensity_match3, var.name = "distance")
```

```{r message = FALSE, warning = FALSE, fig.align='center'}
bal.plot(gam_propensity_match, var.name = "gp_19_20")
bal.plot(gam_propensity_match, var.name = "pts_19_20") # look into
bal.plot(gam_propensity_match, var.name = "ppg_19_20") # look into
bal.plot(gam_propensity_match, var.name = "age_continuous")
bal.plot(gam_propensity_match, var.name = "pm_19_20")
bal.plot(gam_propensity_match, var.name = "position")
```



