---
title: "Does Not Playing Hockey Make You Worse at Hockey? Project Progress 7-21-22"
author: "Jackie Jovanovic, Michele Sezgin"
output: 
  html_document:
    code_folding: hide
date: 'July 21st, 2022'
---

```{r, include = FALSE}
knitr::opts_chunk$set(echo = FALSE, message = FALSE, warning = FALSE)
```


```{r load-packages, message=FALSE}
library(tidyverse)
library(lubridate)
library(ggridges)
# Jackie
library(RColorBrewer)
library(ggbeeswarm)
```

### Purpose

During 2020-2021, many hockey leagues (including NHL feeder leagues) had shortened seasons or no season due to the restrictions on play caused by the COVID-19 pandemic. Some players experiencing restrictions on play in played in other leagues or tournaments during the 2021-2022 season. Others did not play any league/tournament games during the 2020-2021 season. This poses the question of whether not playing games during the 2020-2021 COVID season negatively impacted player development (or caused players to get worse). To answer this question, we will examine data from the Ontario Hockey League, which did not play any games during the 2020-2021 season. Some players from this league chose to play in other leagues or tournaments while others did not.

### Data

```{r data}
ohl <- read_csv("../../sams_ohl_data_request.csv")
ohl_covid <- ohl %>%
  filter(season %in% c("2019-2020","2020-2021", "2021-2022"))
```

```{r}
## Add treatment var:

# Filtering only for players who played more than 10 games (should we combine number of games played across leagues?)
ohl_20_21 <- ohl %>%
  filter(season == "2020-2021", gp > 10)

# Split players up into treatment vs non-treatment
ohl_treatment <- ohl %>%
  filter(season %in% c("2019-2020", "2021-2022"), league == "OHL") %>%
  group_by(player_id, season) %>%
  mutate(ppg = sum(pts)/sum(gp),
         treatment = ifelse(player_id %in% ohl_20_21$player_id, "Played", "Didn't play")
  ) %>%
  ungroup()

plyr_quality <- ohl %>%
  filter(season == "2019-2020", league == "OHL") %>%
  group_by(player_id) %>%
  mutate(plyr_quality = sum(pts)/sum(gp)) %>%
  filter(duplicated(player_id) == FALSE) %>%
  ungroup() %>%
  select(player_id, plyr_quality)

# View(plyr_quality %>%
#   filter(duplicated(plyr_quality$player_id) == TRUE))
# ## need to deal with duplicate players/rows
# View(ohl_filtered %>%
#   filter(duplicated(ohl_filtered$player_id) == TRUE))

ohl_pm <- ohl %>%
  filter(season %in% c("2019-2020", "2021-2022"), league == "OHL") %>%
  group_by(team_name, season) %>%
  arrange(team_name, season, desc(pm)) %>%
  mutate(pm_rank = 1:n(),
         pm_relative = pm - mean(pm)
  ) %>%
  ungroup()

# We need to check that each player played in the OHL in 2019-2020 and 2021-2022
ohl_szn <- ohl_pm %>%
  group_by(player_id) %>%
  mutate(played_2019 = season == "2019-2020",
         played_2021 = season == "2021-2022",
         played_both_szn = sum(played_2019) & sum(played_2021)) %>%
  ungroup() %>%
  filter(played_both_szn == TRUE)

# add treatment variable through join
ohl_trt <- ohl_szn %>%
  left_join(select(ohl_treatment, player_id, season, team_name, treatment), by = c("player_id", "team_name", "season")) %>%
  mutate(ppg = pts/gp) 

# add age variable
ohl_age <- ohl_trt %>%
  group_by(player_id, season) %>%
  mutate(year = strsplit(season, "-")[[1]][1]) %>%
  mutate(age = trunc((dob %--% as.Date(paste0(year, "-9-15"))) / years(1)),
         age_continuous = (dob %--% as.Date(paste0(year, "-09-15"))) / years(1)
  ) %>%
  ungroup() #%>%
  #filter(age %in% 16:22)
    
# add player quality
ohl_qlty <- ohl_age %>%
  left_join(plyr_quality, by = "player_id") %>%
  group_by(player_id, season) %>%
  mutate(ppg_total = sum(pts)/sum(gp),
         gp_total = sum(gp),
         pts_total = sum(pts)) %>%
  arrange(player_id, season, -plyr_quality) %>%
  filter(duplicated(player_id) == FALSE) %>%
  ungroup()#%>%
  # filter(ppg != 0) %>%
  # mutate(log_ppg = log(ppg))

# ohl_qlty <- ohl_age %>%
#   left_join(plyr_quality, by = "player_id") %>%
#   group_by(player_id, season) %>%
#   summarize_at(vars(gp, pts), "sum") %>%
#   group_by(player_id) %>% 
#   mutate(ppg_total = pts/gp, count_seasons = n()) %>% 
#   filter(count_seasons == 2) %>%
#   ungroup()

# were they drafted
ohl_filtered <- ohl_qlty %>%
  mutate(drafted = !is.na(draft_year)) %>%
  select(season, player_id, treatment, first_name, last_name, position, plyr_quality, age, gp_total, pts_total, ppg_total, drafted, draft_year, round, overall_pick_num, age_continuous, pm_relative, pm_rank, pm)

## Trying to combine players who played for two teams into one

# combined <- ohl_filtered %>%
  

ohl_treatment_21_22 <- ohl_filtered %>%
  filter(season == "2021-2022")
```

### What proportion of players played during COVID season?

```{r}
#played vs didn't play
num_played <- nrow(filter(ohl_filtered, treatment == "Played"))
num_didnt <- nrow(filter(ohl_filtered, treatment == "Didn't play"))
print(paste0("proportion who played is ", round(num_played/nrow(ohl_filtered), 3)))
```

### EDA (Treatment)

```{r}
# drafted vs treatment
mosaicplot(table("Drafted" = ohl_filtered$drafted,
                 "Treatment" = ohl_filtered$treatment),
           main = "Drafted players more likely to have played during COVID",
           shade = FALSE)
```
#### Continuous Age vs PPG

```{r}
#continuous age vs ppg
ggplot(ohl_filtered, aes(x = age_continuous, y = ppg_total)) +
  geom_point(alpha = .5) +
  labs(title = "No relationship between continuous age and PPG",
       x = "Age (continuous)",
       y = "PPG") +
  theme_bw()
ggplot(ohl_filtered, aes(x = age_continuous, y = ppg_total)) +
  geom_point(alpha = .5, size = .4) +
  facet_grid(rows = vars(age)) +
  labs(title = "No observable pattern in continuous age",
       x = "Age (continuous)",
       y = "PPG") +
  theme_bw()
```

```{r}
# ggplot(ohl_filtered, aes(x = pm_relative, color = treatment)) +
#   geom_density() +
#   labs(title = "Players with better relative PM more likely to have played during COVID?")
# ggplot(ohl_filtered, aes(x = pm_rank, fill = treatment)) +
#   geom_bar() +
#   labs()
```

```{r}
ggplot(filter(ohl_filtered, season == "2019-2020"), aes(x = gp_total, color = treatment)) +
  geom_density() +
  labs("Players who played more games in 2019-2020 more likely to have played during COVID?",
       x = "GP in 2019-2020")
```



### Refit Regression Models

#### MLR with interactions that make sense

```{r code-chunk-label}
mlr_interaction <- lm(ppg_total ~ position*plyr_quality + treatment*drafted + gp_total + plyr_quality*gp_total + plyr_quality*drafted + age_continuous*plyr_quality + season, data = ohl_filtered)
plot(mlr_interaction, which = c(1, 2))
summary(mlr_interaction)
```

#### Transfrom PPG -> log(1 + PPG)

```{r}
mlr_log <- lm(log(1 + ppg_total) ~ position*plyr_quality + treatment*drafted + gp_total + plyr_quality*gp_total + plyr_quality*drafted + age_continuous*plyr_quality + season, data = ohl_filtered)
plot(mlr_log, which = c(1, 2))
summary(mlr_log)
```

Takeaways:

- Does not improve upon previous model modeling PPG, maybe even worse.

#### Transform PPG -> sqrt(1 + PPG)

```{r}
mlr_sqrt <- lm(sqrt(1 + ppg_total) ~ position*plyr_quality + treatment*drafted + gp_total + plyr_quality*gp_total + plyr_quality*drafted + age_continuous*plyr_quality + season, data = ohl_filtered)
plot(mlr_sqrt, which = c(1, 2))
summary(mlr_sqrt)
```

Takeaways:

- Does not improve upon previous model modeling PPG, maybe even worse.

### Non-interaction GLM Gamma inverse link

```{r}
# glm gamma distribution
# set.seed(2004)
ohl_filtered <- ohl_filtered %>%
  mutate(is_train = sample(rep(0:1, length.out = nrow(ohl_filtered))),
         ppg_alt = ppg_total + .001)

ohl_glm <- glm(ppg_alt ~ position + pm_relative + treatment + drafted + gp_total + age_continuous + season,
                         data = ohl_filtered,
                         family = Gamma)
plot(ohl_glm, which = c(1, 2))
summary(ohl_glm)

ohl_filtered <- ohl_filtered %>%
  mutate(glm_resids = ohl_glm$residuals)

ggplot(ohl_filtered, aes(x = pm_relative, y = glm_resids)) +
  geom_point()
ggplot(ohl_filtered, aes(x = age_continuous, y = glm_resids)) +
  geom_point()
```

Takeaways:

- This model does not adequately capture the relationship between our predictors and response.
- **Note**: Tried to fit a GLM with interaction using the default Gamma link function (inverse), but it threw an error and said "no valid set of coefficients has been found: please supply starting values"

```{r}
ohl_glm_interaction <- glm(ppg_alt ~ position*pm_relative + treatment*drafted + gp_total + pm_relative*gp_total + pm_relative*drafted + age_continuous*pm_relative + season, 
                          data = ohl_filtered, 
                          family = Gamma)
 plot(ohl_glm_interaction, which = c(1, 2))
 summary(ohl_glm_interaction)
 
# ohl_filtered <- ohl_filtered %>%
#   mutate(glm_int_resids = ohl_glm_interaction$residuals)
# ggplot(ohl_filtered, aes(x = age_continuous, y = glm_int_resids)) +
#   geom_point()
```

#### Interaction GLM using Gamma distribution with log link function

```{r} 
ohl_glm_interaction <- glm(ppg_alt ~ position*plyr_quality + treatment*drafted + gp_total + plyr_quality*gp_total + plyr_quality*drafted + age_continuous*plyr_quality + season,
                         data = ohl_filtered,
                         family = Gamma(link = log))
plot(ohl_glm_interaction, which = c(1, 2))
summary(ohl_glm_interaction)
```

Takeaways:

- This model does not adequately capture the relationship between our predictors and response (relationship appears non-linear).
- Predictions are not on the correct scale either.

#### Interaction GLM with gaussian log link function

```{r}
# GLM with gaussian log link
ohl_glm_interaction <- glm(ppg_alt ~ position*plyr_quality + treatment*drafted + gp_total + plyr_quality*gp_total + plyr_quality*drafted + age_continuous*plyr_quality + season, 
                         data = ohl_filtered, 
                         family = gaussian(link = log))
plot(ohl_glm_interaction, which = c(1, 2))
summary(ohl_glm_interaction)
```

Takeaways:

- Non-constant variance, not normally distributed towards lower predicted values
- Models relationship better than previous models
- Not predicting PPG on the right scale

### Questions/Concerns

- Next steps for modeling?
- Are we using BART to get propensity scores?


...

