---
title: "Does Not Playing Hockey Make You Worse at Hockey? Project Progress 7-19-22"
author: "Jackie Jovanovic, Michele Sezgin"
output: 
  html_document:
    code_folding: hide
date: 'July 17th, 2022'
---

```{r, include = FALSE}
knitr::opts_chunk$set(echo = FALSE, message = FALSE, warning = FALSE)
```


```{r load-packages, message=FALSE}
library(tidyverse)
library(lubridate)
library(ggridges)
```

### Purpose

During 2020-2021, many hockey leagues (including NHL feeder leagues) had shortened seasons or no season due to the restrictions on play caused by the COVID-19 pandemic. Some players experiencing restrictions on play in played in other leagues or tournaments during the 2021-2022 season. Others did not play any league/tournament games during the 2020-2021 season. This poses the question of whether not playing games during the 2020-2021 COVID season negatively impacted player development (or caused players to get worse). To answer this question, we will examine data from the Ontario Hockey League, which did not play any games during the 2020-2021 season. Some players from this league chose to play in other leagues or tournaments while others did not.

### Data

```{r data}
ohl <- read_csv("../sams_ohl_data_request.csv")
ohl_covid <- ohl %>%
  filter(season %in% c("2019-2020","2020-2021", "2021-2022"))
```

```{r}
## Add treatment var:

# Filtering only for players who played more than 10 games (should we combine number of games played across leagues?)
ohl_20_21 <- ohl %>%
  filter(season == "2020-2021", gp > 10)

# Split players up into treatment vs non-treatment
ohl_treatment <- ohl %>%
  filter(season %in% c("2019-2020", "2021-2022"), league == "OHL") %>%
  group_by(player_id, season) %>%
  mutate(ppg = sum(pts)/sum(gp),
         treatment = ifelse(player_id %in% ohl_20_21$player_id, "Played", "Didn't play")
  ) %>%
  ungroup()

plyr_quality <- ohl %>%
  filter(season == "2019-2020", league == "OHL") %>%
  group_by(player_id) %>%
  mutate(plyr_quality = sum(pts)/sum(gp)) %>%
  filter(duplicated(player_id) == FALSE) %>%
  ungroup() %>%
  select(player_id, plyr_quality)

# View(plyr_quality %>%
#   filter(duplicated(plyr_quality$player_id) == TRUE))
# ## need to deal with duplicate players/rows
# View(ohl_filtered %>%
#   filter(duplicated(ohl_filtered$player_id) == TRUE))


# We need to check that each player played in the OHL in 2019-2020 and 2021-2022
ohl_szn <- ohl %>%
  filter(season %in% c("2019-2020", "2021-2022"), league == "OHL") %>%
  group_by(player_id) %>%
  mutate(played_2019 = season == "2019-2020",
         played_2021 = season == "2021-2022",
         played_both_szn = sum(played_2019) & sum(played_2021)) %>%
  ungroup() %>%
  filter(played_both_szn == TRUE) 

# add treatment variable through join
ohl_trt <- ohl_szn %>%
  left_join(select(ohl_treatment, player_id, season, team_name, treatment), by = c("player_id", "team_name", "season")) %>%
  mutate(ppg = pts/gp) 

# add age variable
ohl_age <- ohl_trt %>%
  group_by(player_id, season) %>%
  mutate(year = strsplit(season, "-")[[1]][2]) %>%
  mutate(age = trunc((dob %--% as.Date(paste0(year, "-05-01"))) / years(1))) %>%
  ungroup() #%>%
  #filter(age %in% 16:22)
    
# add player quality
ohl_qlty <- ohl_age %>%
  left_join(plyr_quality, by = "player_id") %>%
  group_by(player_id, season) %>%
  mutate(ppg_total = sum(pts)/sum(gp),
         gp_total = sum(gp),
         pts_total = sum(pts)) %>%
  arrange(player_id, season, -plyr_quality) %>%
  filter(duplicated(player_id) == FALSE) %>%
  ungroup()#%>%
  # filter(ppg != 0) %>%
  # mutate(log_ppg = log(ppg))

# ohl_qlty <- ohl_age %>%
#   left_join(plyr_quality, by = "player_id") %>%
#   group_by(player_id, season) %>%
#   summarize_at(vars(gp, pts), "sum") %>%
#   group_by(player_id) %>% 
#   mutate(ppg_total = pts/gp, count_seasons = n()) %>% 
#   filter(count_seasons == 2) %>%
#   ungroup()

# were they drafted
ohl_filtered <- ohl_qlty %>%
  mutate(drafted = !is.na(draft_year)) %>%
  select(season, player_id, treatment, first_name, last_name, position, plyr_quality, age, gp_total, pts_total, ppg_total, drafted, draft_year, round, overall_pick_num, shoots)

## Trying to combine players who played for two teams into one

# combined <- ohl_filtered %>%
  

ohl_treatment_21_22 <- ohl_filtered %>%
  filter(season == "2021-2022")
```

#### Wrangling:

**Filtering:**

- 2019-2020 ("pre-COVID") and 2021-2022 ("post-COVID") seasons

- league == OHL

- Only players who played in the OHL during both pre- and post-COVID seasons

**Variables added:**

- points per game per season (combined if a player played for multiple teams in a season)

- games played per season (combined if a player played for multiple teams in a season)

- treatment (i.e. whether a player played more than 10 games during the COVID season)

- age (approximately the oldest a player was in a given season)

- player quality approximated by ppg in pre-COVID season

- whether a player was drafted (not totally up to date)

### EDA

#### Did skaters who played during the COVID season perform better than those who didn't?

```{r}
ggplot(ohl_treatment_21_22, aes(x = ppg_total, color = treatment)) +
  geom_density() +
  labs(title = "Skaters who played during COVID season appear to have\nperformed better post-COVID",
       x = "PPG")
```

But is this difference "real"? Aka did not playing during the COVID season cause players to get worse at hockey or can this difference simply be explained by confounding variables?\n

### EDA with variables of interest: PPG, position, age, player quality, treatment, GP, season

check if explanatory variables are correlated with each other and response.

#### Concerns about GP (and GP vs PPG)

```{r}
# distribution of games played
ggplot(ohl_filtered, aes(x = gp_total)) +
  geom_density() +
  labs(title = "Most players played a significant number of games",
       x = "GP") 
```

```{r}
# is gp correlated with ppg?
ggplot(ohl_filtered, aes(x = gp_total, y = ppg_total)) +
  geom_point(alpha = .5) +
  labs(title = "Relationship between PPG and GP", x = "Total GP", y = "Total PPG")
```

Takeaways:

- players who have inflated ppgs with low games played don't seem to be a concern

- There could be some kind of relationship here -> PPG seems to increase with GP


```{r}
# density curve of ppg faceted by position
ggplot(ohl_filtered) +
  geom_density(aes(x = ppg_total, color = season)) +
  #facet_grid(rows = vars(position)) +
  labs(title = "PPG generally higher in 2021-2022 season",
       x = "PPG")
```

Takeaways: 

- Likely because players get more skilled as they get older, and we are only including players who played both pre- and post-COVID

- Something to control for in our model

#### Position

```{r}
# density curve of ppg faceted by position
ggplot(ohl_filtered) +
  geom_density(aes(x = ppg_total, color = position)) +
  labs(title = "PPG generally greater for forwards in 2019-20 and 2021-22 season",
       x = "PPG")
```

Takeaways:

- Forwards score more than defensemen, which is obvious.

- But, this becomes problematic with the way we're measuring player quality...

**Problems with player quality and PPG**

```{r}
ggplot(ohl_filtered, aes(x = plyr_quality, color = position)) +
  geom_density() +
  labs(title = "Forwards more likely to be high quality players using PPG",
       x = "Player Quality")
```

Takeaways:

- Forwards weighted as better players in our model becuase of our biased metric

- Drafted vs not drafted another way to measure quality, but few players are drafted, and it can be an all or nothing way to measure quality.

```{r}
ggplot(ohl_treatment_21_22) +
  geom_density(aes(x = ppg_total, color = treatment)) +
  facet_grid(rows = vars(position)) +
  labs(title = "PPG generally higher for those who played during COVID season\n in 2021-22 season, regardless of position",
       x = "PPG")
```

Takeaways:

- Position alone is not accounting for difference in PPG for treatment vs non-treatment

#### Age

**What's the distribution of age?**

```{r}
# observations per age group
ggplot(ohl_filtered, aes(x = age)) +
  geom_bar() +
  labs("Distribution of Age",
       x = "Age")
```

**Does PPG increase as players get older?**

```{r}
ggplot(ohl_filtered, aes(x = ppg_total, y = as.factor(age))) +
  geom_density_ridges() +
  labs(title = "PPG likely increases with age",
       x = "PPG",
       y = "Age")
```


**Does player quality increase as players get older?**

```{r}
# player quality vs age
ggplot(ohl_filtered, aes(x = plyr_quality)) +
  geom_density_ridges(aes(y = as.factor(age))) +
  labs(title = "Player quality likely increases with age",
       x = "Player Quality",
       y = "Age")
```

**Were older players more likely to play during COVID (since older players are likely higher quality players)?**

```{r}
# age vs treatment
mosaicplot(table("Age" = ohl_filtered$age,
                 "Treatment" = ohl_filtered$treatment),
           main = "Playing during COVID season independent of age",
           shade = TRUE)
```

Takeaways:

- Not really...

**Do older players play more games?**

```{r}
# age vs gp
ggplot(ohl_filtered, aes(x = gp_total, y = as.factor(age))) +
  geom_density_ridges() +
  labs(title = "Older players more likely to play more games",
       x = "GP")
```

#### Player quality

**Did player quality influence whether someone played during COVID season?**

```{r}
ggplot(ohl_filtered, aes(x = plyr_quality, color = treatment)) +
  geom_density() +
  labs(title = "Players who played during COVID season were more likely to be higher quality players",
       x = "Player Quality")
```

Takeaways:

- We should control for this relationship in our model (interaction term in linear model?)
- Note: there are a lot more players who didn't play than played so this may affect our analysis.

### Modeling

#### Basic multiple regression model, no interaction

```{r, warning = TRUE}
# vanilla MLR
mlr <- lm(ppg_total ~ treatment + season + position + plyr_quality + age, data = ohl_filtered)
#mlr_summary <- summary(mlr)
plot(mlr)
```

#### Does putting PPG on a log scale help meet model conditions for inference?

```{r}
# ppg on log scale
log_mlr <- lm(log(ppg_total) ~ treatment + season + position + plyr_quality + age, data = filter(ohl_filtered, ppg_total != 0))
#summary(log_mlr)
plot(log_mlr)
```

No...

#### Does allowing for all interactions between variables help us meet model conditions?

```{r}
mlr_int <- lm(ppg_total ~ treatment*position*season*plyr_quality*age*drafted, data = ohl_filtered)
#summary(mlr_int)
plot(mlr_int)
```

No... we need a more complex model.

### Questions/Concerns

- PPG is likely not adequately measuring player performance (especially for defensemen)
  - Can we get more stats like ice time, shots blocked, shot on goal, turnovers created, etc?
  - NHL data?
- Multiple regression model accounting for all interactions between explanatory variables does not meet conditions for inference.
- Next steps for building models/maybe starting causal inference?
...
