---
title: "Does not playing hockey make you worse at hockey?"
subtitle: "Project progress"
author: "Michele Sezgin"
date: "`r Sys.Date()`"
output: openintro::lab_report
---

```{r load-packages, message=FALSE}
library(tidyverse)
library(openintro)
library(lubridate)
library(DHARMa)
```

```{r setup, include=FALSE}
knitr::opts_chunk$set(message = FALSE, warning = FALSE)
```

```{r code-chunk-label}
# Rethink data

ohl <- read_csv("../sams_ohl_data_request.csv")

# Create df of players who played during covid w total gp and if they played championship
ohl_treatment_prep <- ohl %>%
  filter(season == "2020-2021") %>%
  group_by(player_id) %>%
  arrange(desc(gp)) %>%
  mutate(gp_covid = sum(gp),
         championship = ifelse(league %in% c("WC","WJC-18", "WJC-20"), TRUE, FALSE),
         covid_league = league[1]) %>%
  ungroup() %>%
  dplyr::select(player_id, gp_covid, championship, covid_league) %>%
  filter(duplicated(player_id) == FALSE)

# Compute player quality vars using pm
ohl_pm <- ohl %>%
  filter(season == "2019-2020", league == "OHL") %>%
  group_by(team_name) %>%
  arrange(team_name, desc(pm)) %>%
  mutate(pm_rank = 1:n(),
         pm_relative = pm - mean(pm)
  ) %>%
  ungroup() %>%
  dplyr::select(player_id, team_name, pm_rank, pm_relative)

# We need to check that each player played in the OHL in 2019-2020 and 2021-2022
ohl_szn <- ohl %>%
  filter(season %in% c("2019-2020", "2021-2022"), league == "OHL") %>%
  group_by(player_id) %>%
  mutate(played_2019 = season == "2019-2020",
         played_2021 = season == "2021-2022",
         played_both_szn = sum(played_2019) & sum(played_2021)) %>%
  ungroup() %>%
  filter(played_both_szn == TRUE)
  
ohl_21_22_ppg <- ohl_szn %>%
  filter(season == "2021-2022") %>%
  group_by(player_id) %>%
  filter(duplicated(player_id) == FALSE) %>%
  mutate(ppg_21_22 = sum(pts)/sum(gp),
         gp_21_22 = sum(gp)) %>%
  dplyr::select(player_id, ppg_21_22, gp_21_22)

# Add treatment vars and pm
ohl_treatment <- ohl_szn %>%
  filter(season == "2019-2020") %>%
  left_join(ohl_treatment_prep, by = "player_id") %>%
  left_join(ohl_pm, by = c("player_id", "team_name")) %>%
  replace_na(list(gp_covid = 0, championship = FALSE))

# add age variable
ohl_age <- ohl_treatment %>%
  group_by(player_id) %>%
  mutate(age = trunc((dob %--% as.Date("2020-01-01")) / years(1)),
         age_continuous = (dob %--% as.Date("2020-01-01")) / years(1)
  ) %>%
  ungroup()

# summarizing variables, adding player performance as ppg
ohl_performance <- ohl_age %>%
  group_by(player_id) %>%
  mutate(ppg_19_20 = sum(pts)/sum(gp),
         gp_19_20 = sum(gp),
         pts_19_20 = sum(pts),
         pm_relative_19_20 = mean(pm_relative),
         pm_rank_19_20 = trunc(mean(pm_rank)),
         pm_19_20 = pm,
         treatment = ifelse(gp_covid > 0, "Played", "Didn't Play")
  ) %>%
  arrange(player_id, season) %>%
  filter(duplicated(player_id) == FALSE) %>%
  ungroup()

# were they drafted?
ohl_drafted <- ohl_performance %>%
  mutate(drafted = case_when(draft_year != 2021 ~ TRUE,
                             draft_year == NA ~ FALSE,
                             TRUE ~ FALSE))

# add 2021 ppg
ohl_filtered <- ohl_drafted %>%
  left_join(ohl_21_22_ppg, by = "player_id")

# clean up
ohl_filtered <- ohl_filtered %>%
  dplyr::select(season, player_id, championship, gp_19_20, first_name, last_name, position, age, gp_19_20, pts_19_20, ppg_19_20, drafted, draft_year, round, overall_pick_num, age_continuous, pm_relative_19_20, pm_rank_19_20, pm_19_20, treatment, ppg_21_22, gp_covid, gp_21_22, covid_league)
```

## Exploratory Data Analysis

### Response: Player Performance Post-COVID Season {.tabset}

#### Treatment vs Performance

```{r, fig.align='center', warning=FALSE, message=FALSE}
# treatment vs ppg
ggplot(ohl_filtered, aes(x = ppg_21_22, color = treatment)) +
  geom_density() +
  labs(title = "Players who played during COVID season generally had higher PPG in\n2021-22 season",
       x = "Player Performance (PPG Post-COVID Season)",
       color = "COVID season") +
  scale_color_manual(values = c("royalblue3", "darkgoldenrod3")) +
  theme_bw()
```

#### Age vs Performance

```{r}
# age vs ppg
ggplot(ohl_filtered, aes(x = age_continuous, y = ppg_21_22)) +
  geom_point(alpha = .5) +
  labs(title = "No relationship between player performance and age",
       x = "Age pre-COVID",
       y = "Player Performance (PPG Post-COVID Season)") +
  theme_bw()
```

#### GP vs Performance

```{r}
# gp vs ppg
ggplot(ohl_filtered, aes(x = gp_21_22, y = ppg_21_22)) +
  geom_point(alpha = .5) +
  labs(title = "Positive linear relationship between player performance and GP in \n2021-22 season",
       x = "GP in 2021-22 Season", 
       y = "Player Performance (PPG Post-COVID Season)") +
  theme_bw() +
  geom_smooth(method = "lm")
```

#### Drafted vs Performance

```{r}
# drafted vs ppg
ggplot(ohl_filtered) +
  geom_density(aes(x = ppg_21_22, color = drafted)) +
  labs(title = "Post-covid season player performance generally greater for drafted players",
       x = "Player Performance (PPG Post-COVID Season)",
       color = "Drafted") +
  scale_color_manual(values = c("royalblue3", "darkgoldenrod3")) +
  theme_bw()
```

#### Player Quality vs Performance

```{r}
# ppg 2019-20 vs ppg 21-22
ggplot(ohl_filtered, aes(x = ppg_19_20, y = ppg_21_22)) +
  geom_point(alpha = .5) +
  labs(title = "Positive linear relationship between post-COVID season player performance \nand player quality",
       x = "Player Quality (PPG in pre-COVID Season)",
       y = "Player Performance (PPG Post-COVID Season)") +
  geom_smooth(method = "lm") +
  theme_bw()
```

#### Position vs PPG

```{r}
# position vs ppg
ggplot(ohl_filtered) +
  geom_density(aes(x = ppg_21_22, color = position)) +
  labs(title = "Post-COVID season player performance generally greater for forwards",
       x = "Player Performance (PPG Post-COVID Season)",
       color = "Position") +
  scale_color_manual(values = c("royalblue3", "darkgoldenrod3")) +
  theme_bw()
```

### {-}

### Treatment {.tabset}

#### Drafted vs Treatment

```{r, fig.align='center'}
# drafted status vs treatment
mosaicplot(table("Drafted" = ohl_filtered$drafted,
                 "COVID Season" = ohl_filtered$treatment),
           main = "Drafted players more likely to have played during COVID season",
           shade = TRUE)
```

#### Player Quality vs Treatment

```{r}
# pm vs treatment
ggplot(ohl_filtered, aes(x = ppg_19_20, color = treatment)) +
  geom_density() +
  labs(title = "Better quality players more likely to have played during COVID season", 
       x = "Player Quality (PPG in pre-COVID season)",
       color = "COVID Season") +
  scale_color_manual(values = c("royalblue3", "darkgoldenrod3")) +
  theme_bw()
```

#### Age vs Treatment

```{r}
# age vs treatment
ggplot(ohl_filtered, aes(x = age_continuous, color = treatment)) +
  geom_density() +
  labs(title = "Insignificant difference in age for players who played during COVID season and players who did not", 
       x = "Age",
       color = "COVID Season") +
  scale_color_manual(values = c("royalblue3", "darkgoldenrod3")) +
  theme_bw()
```

#### GP vs Treatment

```{r}
# gp vs treatment
ggplot(ohl_filtered, aes(x = gp_19_20, color = treatment)) +
  geom_density() +
  labs(title = "Insignificant difference in GP during 2019-20 season for players\nwho played during COVID season", 
       x = "GP",
       color = "COVID season") +
  scale_color_manual(values = c("royalblue3", "darkgoldenrod3")) +
  theme_bw()
```

### {-}

### Explanatory Variable Relationships {.tabset}

#### Age vs Player Quality

```{r}
ggplot(ohl_filtered, aes(x = age_continuous, y = ppg_19_20)) +
  geom_point(alpha = .5) +
  labs(title = "Positive linear relationship between player quality and age",
       x = "Age pre-COVID",
       y = "Player Quality (PPG pre-COVID season)")
```


#### Age vs Drafted

```{r}
ggplot(ohl_filtered, aes(x = age_continuous, color = drafted)) +
  geom_density() +
  labs(title = "Drafted players younger",
       x = "Age pre-COVID")
```

#### Position vs Player Quality

```{r}
ggplot(ohl_filtered, aes(x = ppg_19_20, color = position)) +
  geom_density() +
  labs(title = "Forwards more likely to be 'quality' players",
       x = "Player Quality (PPG pre-COVID Season)",
       color = "Position")
```

#### 2019-20 GP vs Player Quality

```{r}
ggplot(ohl_filtered, aes(x = gp_19_20, y = ppg_19_20)) +
  geom_point(alpha = .5) +
  theme_bw() +
  geom_smooth() +
  labs("Player quality increases non-linearly with GP pre-COVID season",
       x = "GP pre-COVID Season",
       y = "Player Quality (PPG pre-COVID Season)")
```

### {-}

## Modeling

### Linear Models {.tabset}

#### Intercept Only (Average)

```{r}
intercept_only <- lm(ppg_21_22 ~ 1, data = ohl_filtered)
summary(intercept_only)
plot(intercept_only, which = c(1, 2))
```


#### Basic OLS

```{r}
ohl_mlr <- lm(ppg_21_22 ~ position + ppg_19_20 + treatment + drafted + gp_19_20 + age_continuous, data = ohl_filtered)
plot(ohl_mlr, which = c(1, 2))
summary(ohl_mlr)
```

##### OLS model not predicting PPG below zero anymore!

```{r}
#min(ohl_mlr$fitted.values)
```

#### Interaction OLS

```{r}
# ohl_mlr_interaction <- lm(ppg_21_22 ~ position*ppg_19_20 + drafted*treatment + gp_19_20*ppg_19_20 + age_continuous*ppg_19_20 + ppg_19_20*treatment + ppg_19_20*drafted + drafted*age_continuous + gp_21_22, data = ohl_filtered)
# plot(ohl_mlr_interaction, which = c(1, 2))
# summary(ohl_mlr_interaction)
```
#### Gamma Regression

```{r}
# ohl_filtered <- ohl_filtered %>%
#   mutate(ppg_alt = ppg_21_22 + .001)
# ohl_glm_interaction <- glm(ppg_alt ~ position*ppg_19_20 + drafted*treatment + gp_19_20*ppg_19_20 + age_continuous*ppg_19_20 + ppg_19_20*treatment + ppg_19_20*drafted + drafted*age_continuous,
#                           data = ohl_filtered, 
#                           family = Gamma)
# plot(ohl_glm_interaction, which = 1)
# summary(ohl_glm_interaction)
# simout  <-  simulateResiduals(ohl_glm_interaction)
# plotSimulatedResiduals(simout) 
```

Conditions not met for gamma regression, residuals not what we would expect under gamma-distribution.

#### Mixed Effects model

```{r}
# ggplot(filter(ohl_filtered, treatment == "Played"), aes(x = ppg_21_22, color = covid_league)) +
#   geom_density()
```


```{r}
ohl_filtered <- ohl_filtered %>%
  mutate(treatment_numeric = ifelse(treatment == "Played", 1, 0))

table(ohl_filtered$covid_league)

library(lme4)
ohl_filtered$new_covid_league <- ifelse(is.na(ohl_filtered$covid_league), "NONE",
                                        ohl_filtered$covid_league)
ohl_lmer <- lmer(ppg_21_22 ~ position + ppg_19_20 + drafted + gp_19_20 + age_continuous + treatment_numeric + (treatment_numeric|new_covid_league),
                     data = ohl_filtered)
summary(ohl_lmer)
# anova(ohl_lmer)
```

```{r}
VarCorr(ohl_lmer) %>% as_tibble() %>% mutate(icc = vcov / sum(vcov)) %>% dplyr::select(grp, icc)
library(merTools)
league_effects <- REsim(ohl_lmer)


league_effects %>%
  as_tibble() %>%
  group_by(groupFctr) %>%
  arrange(desc(mean)) %>%
  slice(1:5, (n() - 4):n()) %>%
  ggplot(aes(x = reorder(groupID, mean))) +
  geom_point(aes(y = mean)) +
  geom_errorbar(aes(ymin = mean - 2 * sd,
                    ymax = mean + 2 * sd)) +
  facet_wrap(~groupFctr, ncol = 1, scales = "free_y") +
  geom_vline(xintercept = 0, linetype = "dashed",
             color = "red") +
  coord_flip() +
  theme_bw() +
  labs(title = "League effect sizes")
plotREsim(league_effects, labs = TRUE)
```


### {-}

#### Full model (no interaction) not significantly different from nested model (interaction)

```{r}
#anova(ohl_mlr, ohl_mlr_interaction)
```

## Questions/Concerns:

1. Should we limit drafted to be drafted draft year 2019 or 2020, excluding 2021? 

1. Do I need to separate age into pre- and post-covid and use 2019 age in the interaction between age and player quality?

...

